---
title: "Phage Tor is a potent, broad-host range candidate for inclusion in cocktails targeting Pseudomonas aeruginosa"
format: 
  docx:
    reference-doc: custom-reference-doc.docx
editor: visual
execute: 
  echo: false
  cache: true
bibliography: references.bib
---

```{r}
#| label: setup
#| include: false
#| cache: false
#| out-width: 8.3in
#| fig-dpi: 300

#Please note this document was rendered using Knitr v1.5.0. Differences in output may be caused by a different rendering engine, such as julia or jupyter, for which this remains untested.

library(remotes)
#remotes::install_version("ggplot2", version = "3.5.2") # need this version to ensure compatability with ggtree and cowplot (urgh)

#References here are crude, full references were added later in Microsoft Word using Zotero. 

# Load relevant packages
library(dplyr) # basic data manipulation
library(tidyr)
library(stringr)
library(ggplot2)
library(purrr)
library(cowplot) # for plotting grids and panels of figure
library(ape) # for reading in .treefiles
library(boot) # for bootstrapping, should it be needed
library(DescTools) # utility tools
library(gt) # for displaying tables, should it be needed
library(magick) # for reading in image files, as many figures are images
library(bipartite) # for analysis of the network for nestedness
library(grid) # to plot a complexheatmap object using cowplot
library(sessioninfo) #summarise which versions of everything I am using
library(ggtext) #to italicise names in axis

# Install ggtree from GitHub if not already installed. Need dev versions to work with ggplot 4.0.1
if (!requireNamespace("ggtree", quietly = TRUE)) {
  remotes::install_github("YuLab-SMU/ggtree")
  remotes::install_github("YuLab-SMU/treeio")
  remotes::install_github("YuLab-SMU/tidytree")
}


library(ggtree)
library(treeio)


#these are the packages needed for pretty much all code chunks. In some cases there are a few others lying about thoughout the code. When rendering, there may be a couple rounds of "this package isnt installed". Bear that in mind! 

set.seed(42) # seed for randomness control
okabe <- c("#8dd3c7", "#bebada", "#fb8072", "#80b1d3", "#fdb462", "#b3de69", "#fccde5", "#d9d9d9", "#a35eb5","#80c683","#8c8cd4", "#bc80bd") # colour palette
```

Christian Fitch^1^, Ellie Tong^1^, Jonathan Crane^1,2^, Thomas Murray^4^, Luis Bolaños^1^, Julie Fletcher^1^, CPL Citizen Scientists, Vicki Gold^1,3^, Eshwar Mahenthiralingam^5^, Stephen Michell^1^, Phillip Mitchelmore^6,7^, Ben Temperton^1,8\*^ 

^1^ Biosciences, Faculty of Health and Life Sciences, University of Exeter, Exeter, UK 

^2^ Natural Sciences, Faculty of Environment, Science, and Economy, University of Exeter, Exeter, UK

^3^ Living Systems Institute, Stocker Road, University of Exeter, Exeter, UK

^4^ Department of Pediatrics, Yale School of Medicine, New Haven, Connecticut, USA 

^5^ School of Biosciences, Cardiff University, Cardiff, UK 

^6^ Department of Respiratory Medicine, Royal Devon University Hospital NHS Foundation Trust, Exeter, UK 

^7^ NIHR Exeter BRC, University of Exeter, Exeter, UK

^8^ GutSee Health Ltd, Quadram Institute, Norwich, UK

**\*Corresponding Author** 

[ben\@gutseehealth.com](mailto:ben@gutseeehealth.com)

Target Journal: [Antimicrobial Agents and Chemotherapy](https://journals.asm.org/journal/aac/article-types#research-articles). No limit on word count, figures, tables, references (although it'll cost more to publish, I think). 250-word abstract. Format neutral journal. 

Running Title: Screening-Based Therapeutic Phage Cocktail Design 

Key Words: Antibiotic Resistance, Phage Therapy, Bacteriophage, Phage Library, Phage Cocktails, Phage Bacteria Interaction Networks, Pseudomonas aeruginosa, Cystic Fibrosis, Bronchiectasis 

Author contributions: **Christian Fitch**: Conceptualization, Data Curation, Formal Analysis, Investigation, Methodology, Resources, Visualisation, Writing - Original Draft, Writing - Review & Editing; **Ellie Tong**: Investigation, Writing - Review & Editing; **Jonathan Crane**: Formal Analysis, Visualisation, Writing - Review & Editing; **Thomas Murray**: Investigation, Methodology, Supervision, Writing - Review & Editing; **Luis Bolaños**: Investigation, Writing - Review & Editing; **Julie Fletcher**: Investigation, Resources, Writing - Review & Editing; **CPL Citizen Scientists**: Resources; **Vicki Gold**: Formal Analysis, Methodology, Supervision, Writing - Review & Editing; **Eshwar Mahenthiralingam**: Resources, Supervision, Writing - Review & Editing; **Stephen Michell**: Supervision, Writing - Review & Editing; **Phillip Mitchelmore**: Resources, Supervision, Writing - Review & Editing; **Ben Temperton**: Conceptualization, Formal Analysis, Funding Acquisition, Methodology, Project Administration, Resources, Supervision, Writing - Review & Editing

Acknowledgements: We are grateful to all citizen scientists who took part in outreach events and collected water samples for the project, as well as Nick Evens and the Environment Agency team for providing wastewater samples. We would also like to acknowledge the Exeter Sequencing Centre at the University of Exeter for preparing and sequencing Illumina libraries for phage short read sequencing utilising equipment funded by the Wellcome Trust (Multi-User Equipment Grant award number 218247/Z/19/Z) and Dr Christian Hacker and Grace Patterson at the University of Exeter Bioimaging Centre for TEM imaging of phages. We are grateful to Dr Steve Porter for training in production of bacterial knockout mutants, Professor Jo Fothergill for supplying *Pseudomonas* phage PNM, and Pamela Huang for support during the acquisition of cascade impactor data. This work was supported in part by grant MR/W006308/1 for the GW4 BIOMED2 DTP, awarded to the Universities of Bath, Bristol, Cardiff, and Exeter from the Medical Research Council (MRC)/UKRI, and an InnovateUK Biomedical Catalyst award (10070793) to BT, and NIHR Exeter BRC to PM. EM acknowledges support for the Cardiff University bacterial strain collection from the UK Cystic Fibrosis Trust Infection Biorespository (CFTP Award No: 0006).

## Abstract

Bacteriophage (phage) therapy is a promising option for the future management of bacterial infections where antibiotics are ineffective. The high specificity of phages to individual pathogens often necessitates phage matching prior to treatment. However, prioritising use of broader host-range phages and/or cocktails can alleviate this, enabling the same phage treatment to be provided to many patients. Such strategies optimise phage therapy from logistic, economic, and scalability perspectives. Here, rapid screening of 37 phages from the Citizen Phage Library identified *Pseudomonas* phage Tor, a *Pbunavirus* that inhibited the growth of 23 of 25 *P. aeruginosa* strains representative of its species diversity. Tor was shown to use the outer-core of the lipopolysaccharide as a receptor; be stable in clinically relevant buffers; compatible with delivery via nebulisation; and adaptable to novel hosts via co-evolution. We demonstrate that Tor is a potent phage around which broad-host range cocktails can be built, with other phages or adjuvants selected to further increase host-range and killing efficacy.

{{< pagebreak >}}

## Introduction

Bacterial resistance to antibiotics presents a major threat to human health. Current data suggests that antibiotic resistance is already an important cause of mortality (murray 2022), and worsening over time (cassini 2019). It has been predicted that by 2050, ten million deaths per year will be attributable to antibiotic resistance (o'neill 2016). This presents a serious global health crisis that needs to be addressed. Reductions in the number of therapeutic options for patients with drug-resistant infections, limited novel antibiotics in development (butler 2023), and individual patient intolerances to current antimicrobials (koliscak, 2013) emphasise the urgent need for alternative treatment options.

Bacteriophage (phage) therapy, the therapeutic use of bacteria-lysing viruses, is increasingly used as an alternative or adjunct to antibiotic therapy. Lytic phages infect and kill bacterial cells as part of their replication cycle, and have been used successfully to treat bacterial infections for over a century (Summers 2012), particularly in Eastern Europe (Kutateladze, 2008). Recent consecutive case series, which address potential reporting biases, also support the utilisation of phage therapy (Young 2023; Pirnay 2024), and a combined analysis of published case studies highlights phages' remarkable safety profile with overall clinical improvement observed in 78.8% of patients (Uyttebroek 2022).

*Pseudomonas aeruginosa* is recognised as an important pathogen for phage therapy development. It is a major contributor to the antimicrobial resistance crisis (Murray, 2022), particularly in the context of chronic lower respiratory infections in patients with conditions such as Cystic Fibrosis and bronchiectasis, where a high burden of anti-pseudomonal therapy is part of routine care. Consequently, these patient cohorts have a high need for alternative treatments. In addition, phages targeting clinical *P. aeruginosa* are numerous within phage banks of characterised phages, such as the Citizen Phage Library (CPL; Fletcher, Tong), and is the most frequently represented pathogen for phage therapy requests, with successful treatment now shown in multiple cases (Aslam 2019; Pirnay 2024; Chan 2025; Weiner 2025).

The strain-level specificity of phages, combined with the diversity of *P. aeruginosa* isolates from clinical settings (Freschi 2018) and even within patients (Hilliam 2017, mitchelmore 2016), often requires matching individual phages from pre-defined panels to patient isolates prior to the initiation of phage therapy (Pirnay 2024, Chan 2025). In addition to higher costs and development burden of personalised treatments, efforts to optimise phage use must overcome several challenges: neutralisation of phage via interactions with the immune system (Van bellegham 2019; Chan 2021); maintaining and manufacturing large and diverse phage biobanks (Fletcher 2024; Humolli 2025); development of bacterial resistance to phages during therapy (Schooley 2017; Burmeister and Turner, 2020; Chan 2025); and challenges in measuring pharmacokinetics during phage treatment (Payne and Jansen 2003; Chow 2020).

Broad host-range phages, which can lyse multiple bacterial strains, offer the potential to reduce the cost and time burden associated with personalised phage therapies. Broad-spectrum phages with complementary host ranges (Abedon 2021), that collectively target 80% of currently circulating bacterial strains (Merabishvilli 2018) have high commercial value. Litre-scale manufacture of phages can produce approximate phage titres of 10^11^ plaque forming units per millilitre (PFU/mL), equivalent to hundreds of doses per batch at a titre of 10^9^ PFU (Fletcher 2024; khatami 2022).

Per-treatment cost of phage products is strongly influenced by economies of scale, phage stability, host-range, and cocktail complexity. If the same phage product can be used for a small number of patients (broad host-range) and demonstrates high stability (long shelf life), then even with costs associated with Good Manufacturing Practice, phage therapies can achieve per-treatment cost parity with modern antibiotics (Iredell and Sacher, 2023). Further increases in stability and host-range translate into more efficient use of phage batches, improving an already favourable cost-benefit analysis for phage therapy with the UK's National Health Service (Scottish Health Technologies Group, 2023).

Identification of individual phages with broad host-ranges typically involves screening a panel of phages against a representative panel of bacterial isolates to produce a phage-bacteria interaction network (PBIN; Weitz 2013). This enables either selection of the broadest phages or the design of broad-spectrum cocktails, either by manual curation or by using computational packages (e.g. PhageCocktail (Menor-Flores 2022). To finalise cocktail composition, phages should be tested for stability (Duyvejonck 2021), lack of antagonism with standard-of-care antibiotics (Kever 2022, liu 2020), and other biological features (genome sequence, burst size, bacterial resistance mechanisms, complementary receptor use) to increase likelihood of therapeutic success. In addition, effective delivery to the intended site is crucial (e.g. nebulisation to the lungs; flint 2023).

Here, we first evaluated the ability of a panel of 37 phages targeting *P. aeruginosa* to inhibit the growth of a panel of genomically diverse bacterial strains. This initial screen identified phage Tor as an extremely broad host-range phage. Further characterisation supported the potential use of the phage in therapy: Tor is stable in relevant delivery buffers, robust to shear forces associated with nebulisation, and can be co-evolved to expand its host-range. Further improvements to killing efficiency when combined with antibiotics and other phages show that Tor serves as an effective basis for designing broad-spectrum phage products.

## Results

#### The Citizen Phage Library (CPL) collection of phages targeting *P. aeruginosa* is narrow

To characterise our phages that show lytic activity against *P. aeruginosa*, we selected 37 phages that represented all sequenced *Pseudomonas*-targeting phages within the CPL at the time of experimentation. These phages, including original isolation hosts, location of origin, citizen-supplied names, genomic information, and predicted receptor are summarised in Table S1. Pairwise intergenomic similarity calculations between all 37 phages revealed these phages belonged to three viral genera (Figure 1A): *Abidjanvirus* (CPL00212, CPL00214, CPL00215, CPL00216, and CPL00217; 13.5% of total), *Yuavirus* (CPL00161, CPL00211, CPL00252, CPL00253, and CPL00254; 13.5% of total), and *Pbunavirus* (remainder; 73.0% of total), and 13 viral species. Phages isolated through the CPL protocols (Fletcher, 2024) were limited in diversity to three clusters across *Pseudomonas* phage diversity within the INPHARED (Cook, 2021) database (Figure 1B).

#### The CPL *P. aeruginosa* host strain panel is representative of global *P. aeruginosa* diversity

The CPL panel of 25 predominantly clinical *P. aeruginosa* isolates were selected for screening phage host range (Table S2; 20 respiratory tract, 1 bloodstream, 3 Diabetic Foot Infection (DFI), 1 lab strain). To evaluate phylogenomic diversity within this test, the CPL panel was compared by core gene phylogenetic analysis with 39 isolates from the International *P. aeruginosa* strain panel (IPAP: Freschi 2018, Table S2), designed in accordance with principles for therapeutic testing (Mahenthiralingam 2022). The majority of isolates resided phylogenetically in *P. aeruginosa* group 1 (68%; Figure 1C; Figure S1), which is the most common lineage within this species and represents 75% of isolates genome sequenced and published up to 2018 (freschi 2018). Most other isolates placed within *P. aeruginosa* phylogenetic group 2 (22.7%; Figure 1C; Figure S1), which is representative of 22% of genome sequenced isolates up to 2018 (freschi 2018). Two of the panel strains (DFI10.1 and PAO258) fell outside of the diversity of the reference panel and most likely belong to phylogenetic groups 3, 5, or 5 as defined in Freschi *et al* 2018.

Overall, our test isolates were distributed across the diversity of the IPAP (Figure 1C; Supplementary Table S2). This shows that the CPL *P. aeruginosa* collection was representative of clinical *P. aeruginosa* diversity as observed in previous genomic studies (Freschi 2018) and is suitable for therapeutic testing (Mahenthiralingam 2022) of the *P. aeruginosa* phage collection.

{{< pagebreak >}}

```{r fig-fig1genomics, fig.asp=2, fig.width=24, warning=FALSE, message=FALSE, dpi=300, out.height='10in', out.width='5in'}
#| fig-cap: "Genetic characterisation of the CPL phage and P. aeruginosa panels. A) VIRIDIC intergenomic similarity heatmap of sequenced phages from within the CPL. B) vConTACT3 network of phage similarity. Grey nodes refer to Pseudomonas phages within the INPHARED database, and CPL phages are coloured as blue (Pbunaviruses), red (Abidjanviruses), and orange (Yuaviruses) nodes. C) Rectangular cladogram representing core gene (n = 2745 core genes) phylogeny of CPL hosts (purple text) alongside genomes of the International P. aeruginosa reference panel (orange text). Confidence values are only shown before nodes with less than 90% confidence"
#| fig-dpi: 300

#figure1A-----------------------------------------------------------------------
#read in the data output from VIRIDIC
phage_genomes <- read.csv("rawdata/figure1A/Raw data about each phage no 56.csv") %>%
  pivot_longer(!genome, names_to = "comparison", values_to = "similarity") %>%
  mutate(genome = ifelse(genome == "CPL00272", "CPL00272 (Tor)", genome)) %>%
  mutate(comparison = ifelse(comparison == "CPL00272", "CPL00272 (Tor)", comparison))

#isolate the order of the VIRIDIC output to ensure final graph is the same order (matching VIRIDIC's clustering)
phage_order <- phage_genomes %>%
  select(genome) %>%
  unique() %>%
  as.vector() %>%
  unlist()

#set the order of the phage genomes for ggplot
phage_genomes <- phage_genomes %>%
  mutate(genome = factor(genome, levels = phage_order)) %>%
  mutate(comparison = factor(comparison, levels = phage_order))
  

#Make the plot
phage_genome_plot <- ggplot(phage_genomes, aes(x = genome, y = comparison, fill = similarity)) +
  geom_tile() +
  scale_y_discrete("Phage Genome") +
  scale_x_discrete("Phage Genome") +
  scale_fill_gradient2("Whole Genome\nIntergenomic\nSimilarity (%)", limits = c(0,100), low = "#ffffff", mid = "#d8d7f5", midpoint = 60, high = "#0400ff") +
  theme_cowplot(32) +
  theme(axis.text.x = element_text(angle = 90))

#figure1B-----------------------------------------------------------------------
#need to read in the vcontact3 data
# use magick to read in the png produced by vcontact, cytoscape, then inkscape
network_image <- magick::image_read("rawdata/figure1B/Cytoscape Work/final_figure.png")
#too big, have to resise
network_image <- magick::image_scale(network_image, "4000")
#now make the figure
figure1b <- ggplot() + theme_void() + draw_image(network_image)


#figure1C-----------------------------------------------------------------------

#setup
nodeid.tbl_tree <- utils::getFromNamespace("nodeid.tbl_tree", "tidytree")
rootnode.tbl_tree <- utils::getFromNamespace("rootnode.tbl_tree", "tidytree")
offspring.tbl_tree <- utils::getFromNamespace("offspring.tbl_tree", "tidytree")
offspring.tbl_tree_item <- utils::getFromNamespace(".offspring.tbl_tree_item", "tidytree")
child.tbl_tree <- utils::getFromNamespace("child.tbl_tree", "tidytree")
parent.tbl_tree <- utils::getFromNamespace("parent.tbl_tree", "tidytree")

#read in newick file produced using roary (script in rawdata/figure1C folder) and a csv of which strain originated where
nwk <- read.newick("rawdata/figure1C/core-genes-full.nwk", node.label='support')
df_tip_data <- read.csv("rawdata/figure1C/panel_strains.csv") %>% 
  mutate(Origin = ifelse(panel == "Y", "Panel Strain", "This Study")) %>% 
  select(-panel)

#plot it according to ggtree documentation
p1 <- ggtree(nwk, branch.length = "none") %<+% df_tip_data
bacterial_genome_plot <- ggtree::rotate(p1, 64) +
  geom_tiplab(size=11, aes(color=Origin)) +
  scale_colour_manual(values = c("#fdb462", "#a35eb5")) +
  #geom_nodepoint(color="black", size=0.1) +
  geom_nodelab(geom='label', aes(label=support, subset=support < 0.9), hjust = 1.4, size = 10) +
  geom_hilight(node=69, fill="steelblue", alpha=.4) +
  geom_hilight(node=c(109,85, 82, 83), fill="darkgreen", to.bottom = TRUE, alpha=.4) +
  hexpand(.04, direction = 1) +
  theme(legend.position = c(0.1,0.9),
        legend.text = element_text(size = 32),
        legend.title = element_text(size = 32))

#end---------------------------------------------------------------------------

figure <- plot_grid(phage_genome_plot, figure1b, bacterial_genome_plot, labels = "AUTO", nrow = 3, rel_heights = c(2.6,1.5,3.6), label_size = 28)
figure
ggsave("PDF-of-figures/figure1.pdf", plot = figure, dpi = 300)
```

{{< pagebreak >}}

#### Pairwise phage screens maximise the likelihood of selecting the broadest phages

We first evaluated whether phage with highly similar phenotypes interacted similarly with bacterial hosts. CPL00272 (subsequently named Tor) and CPL00950 - two phages with 94.09% intergenomic similarity and shared genomic architecture were selected (Figure 1A, Figure S2). Tor and CPL00950 both require the outer core region of lipopolysaccharide (LPS) to initiate infection of bacterial hosts, as evidenced by loss of infection in outer core LPS knockout mutants (Figure 2A). They also generated cross-resistance to each other (Figure S3). Despite these similarities between the phages, Tor has a broader host range, and was able to reduce endpoint optical density at 24 hours by over 50% for 60% of hosts, compared to 44% of hosts for CPL00950 (Figure 2B; Figure S4). Furthermore, phage CPL00071 is 92.9% genomically similar to Tor and resides in the same cross resistance group (Kim 2024; Figure S3; Figure S5), despite requiring a different region of the LPS for infection (O-antigen; Tong 2026), showing that selection of receptor targets does not necessarily predict cross resistance patterns needed for robust cocktail design.

A pairwise phage-bacteria interaction network (PBIN; Figure 2C) was produced in order to identify the broadest, and thus most useful phages to take forward for broad-spectrum use. As with Tor and CPL00950, differences in host range between highly similar phages were common, CPL00283 was able to inhibit the growth of 21/25 hosts, but CPL00075 only inhibited the growth of 8/25 hosts, despite 94.4% intergenomic similarity and both being predicted to use the outer core LPS (Tong, 2026) as a receptor.

The limited increases in breadth by additional phages in fully nested PBINs implicates an individual, most-useful phage to take forward for further study and include in prospective cocktail designs or phage panels. Our PBIN was significantly nested (Figure S6; *p* \< 0.001), with the most potent phage (Tor; 23/25 hosts) selected for further study and use in prospective cocktail designs or phage panels. Therefore, pre-selection of phages based on genomic or phenotypic features (e.g. receptor) carries a risk of randomly selecting less-efficient phages (e.g. CPL00950 or CPL00071 instead of Tor).

{{< pagebreak >}}

```{r fig-fig2selection, fig.asp=1.41, fig.width=24, warning=FALSE, message=FALSE, dpi=300, out.height='14.1in', out.width='10in'}
#| fig-cap: "Tor is selected as the best candidate from the CPL. A) Phage titre of three independently diluted stocks of Tor (top) and CPL00950 (bottom) on each of the PAO1 parent strain and clean gene knockouts in PilA (type IV pilus knockout), WbpL (O-antigen), and GalU (outer core LPS). B) Breadth of Tor and CPL00950 assessed by phage-mediated optical density reductions over 50% at 24 hours on the CPL bacterial panel. C) PBIN of 37 tested phages on 25 P. aeruginosa host strains, with inhibition assumed if over than 50% reduction in the area under the untreated bacterial control was observed. Phages are ordered by intergenomic similarity as determined by VIRIDIC above, and bacteria clustered by promiscuity." 
#| fig-dpi: 300

#figure2A-----------------------------------------------------------------------
#receptor data

#read in data, keep only strains of interest, rename the phages to be consistent with the whole story. Remove a cocktail of the two as not relevant.
df = read.csv("rawdata/figure2A/christian cocktail phage pfus.csv") %>%
  mutate(phage = factor(phage, levels = c("P272","P950","cocktail"))) %>%
  filter(bacterial_strain == "PAO1.hsdr" |  bacterial_strain == "PAO1.pilA" |  bacterial_strain == "PAO1.wbpL" | bacterial_strain == "PAO1.galU") %>% 
  filter(phage != "cocktail") %>%
  mutate(phage = ifelse(phage == "P272", "CPL00272 (Tor)", "CPL00950"))

#calculate average phage titres. follow a similar pattern to the section above.
df_average <- df %>% 
  mutate(bacterial_strain = factor(bacterial_strain, levels = c("PAO1.hsdr", "PAO1.pilA", "PAO1.wbpL", "PAO1.galU"))) %>%
  mutate(phage = factor(phage, levels = c("CPL00272 (Tor)","CPL00950"))) %>%
  group_by(bacterial_strain, phage) %>%
  reframe(av_PFU = mean(pfu), stdev = sd(pfu)) %>%
  filter(phage != "cocktail")

#plot this data 
fig2A_top <- ggplot(filter(df_average, phage == "CPL00272 (Tor)"), aes(x=bacterial_strain, y=log10(av_PFU+1))) +
  geom_bar(stat = "identity", alpha = 0.8, fill = "#fdb462") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05)), name = "CPL00272 (Tor) titre\n(Log10(PFU/mL))", breaks = seq(0, 13, 1)) +
scale_x_discrete(labels = c("Wild Type", 
                            "PAO1Δ&thinsp;<i>  pilA</i>", 
                            "PAO1Δ&thinsp;<i>  wbpL</i>",
                            "PAO1Δ&thinsp;<i>  galU</i>")) +
  geom_jitter(data = filter(df, phage == "CPL00272 (Tor)"), aes(x=bacterial_strain, y=log10(pfu+1)), width = 0.01) +
  theme_cowplot(26) +
  xlab("PAO1 Receptor Deletion Mutant") +
  theme(panel.border = element_rect(colour = "black", fill=NA, linewidth=0.5),
        axis.text.x = element_markdown())

fig2A_bottom <- ggplot(filter(df_average, phage == "CPL00950"), aes(x=bacterial_strain, y=log10(av_PFU+1))) +
  geom_bar(stat = "identity", alpha = 0.8, fill = "#80b1d3") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05)), name = "CPL00950 titre\n(Log10(PFU/mL))", breaks = seq(0, 13, 1)) +
scale_x_discrete(labels = c("Wild Type", 
                            "PAO1Δ&thinsp;<i>  pilA</i>", 
                            "PAO1Δ&thinsp;<i>  wbpL</i>",
                            "PAO1Δ&thinsp;<i>  galU</i>")) +
  geom_jitter(data = filter(df, phage == "CPL00950"), aes(x=bacterial_strain, y=log10(pfu+1)), width = 0.01) +
  theme_cowplot(26) +
  xlab("PAO1 Receptor Deletion Mutant") +
  theme(panel.border = element_rect(colour = "black", fill=NA, linewidth=0.5),
        axis.text.x = element_markdown())

#now to do the stats on this section. Data read in identically to above
#only change is making the bacteria a factor to ensure the WT is the baseline
df <- read.csv("rawdata/figure2A/christian cocktail phage pfus.csv") %>%
  mutate(phage = factor(phage, levels = c("P272","P950","cocktail"))) %>%
  filter(bacterial_strain == "PAO1.hsdr" |  bacterial_strain == "PAO1.pilA" |  bacterial_strain == "PAO1.wbpL" | bacterial_strain == "PAO1.galU") %>% 
  filter(phage != "cocktail") %>%
  mutate(phage = ifelse(phage == "P272", "CPL00272 (Tor)", "CPL00950")) %>%
  mutate(bacterial_strain = factor(bacterial_strain, levels = c("PAO1.hsdr","PAO1.pilA", "PAO1.wbpL", "PAO1.galU"))) %>%
  mutate(phage = factor(phage, levels = c("CPL00272 (Tor)", "CPL00950")))

#now can run a linear regression
model_receptor <- lm(log10(pfu+1e-12)~bacterial_strain+phage, data = df)
#plot(model_receptor) # plots are a bit weird, but they are always going to be skewed by the fact that there are similar values and then zeros in the data

#summary(model_receptor) # commented out as do not want this table in the manuscript.

#the model showed a signifcant result. There was a significant difference in the phages - likely due to a slight difference in how phage stocks were adjusted (p = 0.04125), but the differences of imporance are in the receptor knockouts. PilA was significantly lower titre (p = 0.00181) indicating some interaction and GalU was significantly lower (p < 0.001). Outer core LPS therefore required for infection, but a significant role for the type IV pilus.


#figure2b-----------------------------------------------------------------------
#breadth
#Boring beginning things to get my data in 
dictionary <- read.csv("rawdata/figure2B/Dictionary.csv")
strains <- read.csv("rawdata/figure2B/strains.csv")

#okay, so now creating the reformat function to do these things
reformat <- function(df){
  df <- df %>%
    select(-Content) %>%
    slice(-1) %>%
    left_join(dictionary)
  colnames(df)[2] <- "OD600"
  return(df)
}
#now need to make the function that reads in the files in the current working directory
#and adds the plate number, and the time.
produce_df <- function(file_name, time){
  df <- read.csv(file_name, header = TRUE, row.names = NULL)
  df <- reformat(df)
  file_base_name <- sub(".*(Plate \\d+).*", "\\1", basename(file_name))
  df$plate <- file_base_name
  df$time_hours <- time
  df <- df %>% left_join(strains)
  return(df)
}
#Now to go through this for timepoint 0 to make sure I can do it!
csv_files <- list.files(
  path = "rawdata/figure2B/0 hour/", 
  pattern = "\\.CSV$", 
  full.names = TRUE
)
#use purrr::map_df to apply the reformat_csv function to each CSV file and bind them together
df_0 <- map2(csv_files, c(0), produce_df)

#repeat for the other timepoints and then cbind them together
#t19
csv_files <- list.files(
  path = "rawdata/figure2B/19 hour/", 
  pattern = "\\.CSV$", 
  full.names = TRUE
)
#use purrr::map_df to apply the reformat_csv function to each CSV file and bind them together
df_19 <- map2(csv_files, c(19), produce_df)


#t24
csv_files <- list.files(
  path = "rawdata/figure2B/24 hour/", 
  pattern = "\\.CSV$", 
  full.names = TRUE
)
#use purrr::map_df to apply the reformat_csv function to each CSV file and bind them together
df_24 <- map2(csv_files, c(24), produce_df)


#t43
csv_files <- list.files(
  path = "rawdata/figure2B/43 hour/", 
  pattern = "\\.CSV$", 
  full.names = TRUE
)
#use purrr::map_df to apply the reformat_csv function to each CSV file and bind them together
df_43 <- map2(csv_files, c(43), produce_df)

#t48
csv_files <- list.files(
  path = "rawdata/figure2B/48 hour/", 
  pattern = "\\.CSV$", 
  full.names = TRUE
)
#use purrr::map_df to apply the reformat_csv function to each CSV file and bind them together
df_48 <- map2(csv_files, c(48), produce_df)


#produce a df_all for all the data
df_produce <- function(df_list){
  final_df <- Reduce(full_join, df_list)
  return(final_df)
}
#bind them all together
df_0 <- df_produce(df_0)
df_19 <- df_produce(df_19)
df_24 <- df_produce(df_24)
df_43 <- df_produce(df_43)
df_48 <- df_produce(df_48)
df_all <- rbind(df_0,df_19,df_24,df_43,df_48)
#remove the annoying X column
df_all <- df_all %>% select(-X)
#we are not interested in a cocktail between the two, although I have that data, its not relevant anymore
df_all <- df_all %>% filter(Phage != "Cocktail")
#now as I messed up and added phage to B07 to B11, remove these! It also 
#turns out the OD600 row is a character, changing that to numeric!
dodgy <- c("B06", "B07", "B08", "B09", "B10", "B11")
df_all <- df_all %>%
  filter(!(Well %in% dodgy)) %>%
  mutate(OD600 = as.numeric(OD600))

#check the blanks for each plate!
# ggplot(filter(df_all, bacteria == "LB"), aes(x = time_hours, y = OD600, colour = plate)) +
#   geom_point()
#it seems that hour 48 is weird in terms of growth, will remove this data! This is likely due to condensation build up on the plate for some wells. Other timepoint blanks look okay
df_all <- df_all %>%
  filter(time_hours != 48)


#now we want to identify the blank dataframe
df_blanks <- df_all %>%
  filter(bacteria == "LB") %>%
  group_by(time_hours) %>%
  reframe(mean_blk_od = mean(OD600))
#now we will normalise all the ODs
df_all <- df_all %>%
  left_join(df_blanks) %>%
  mutate(norm_od = OD600 - mean_blk_od)
#calculate summary statistics for each of these
df_all_means <- df_all %>%
  group_by(bacteria, Phage, Concentration, time_hours) %>%
  mutate(mean_od_timepoint = mean(norm_od), stdev = sd(norm_od))
#fixing some buggy things regarding the titre being in form 6e+001 or something
df_all <- df_all %>%
  mutate(titre = log10(Concentration)) %>%
  mutate(titre = ifelse(is.infinite(titre) == TRUE, 0, titre)) %>%
  mutate(titre = factor(titre))
#isolate the controls (no phage) and calculate summary statistics
df_controls_only <- df_all %>%
  filter(Phage == "Control") %>%
  group_by(time_hours, bacteria) %>%
  reframe(mean_od = mean(norm_od), stdev = sd(norm_od)) %>%
  mutate(titre = 0)
#remove the control form the original dataframe, as these are clarified elsewhere
df_all <- df_all %>%
  filter(Phage != "Control") 

#to get raw numbers of reduction percentage of the phage at each timepoint we can left join in a means column from df_means
#then calculate reduction pecentage by each phage at each timepoint of interest:

df_reduction <- df_all %>%
  left_join(select(df_controls_only, time_hours, bacteria, mean_od)) %>%
  rename(mean_control_od = mean_od) %>%
  mutate(red_perc = 100*(mean_control_od-norm_od)/mean_control_od)

df_reduction_plot <- df_reduction %>%
  group_by(time_hours, Phage, bacteria, titre) %>%
  reframe(mean_red_perc = mean(red_perc), stdev = sd(red_perc))

df_reduction_plot_modified <- df_reduction_plot %>%
  filter(bacteria != "LB") #remove LB as this one is too noisy for this to make sense

#at some timepoints, the bacterial growth is too low to actually create a meaningful comparison. 
#These bacteria are PIB63.5, PIB50.3, PIC1.4, PIB52.4, RDE3, RDE4 cause particular issues with this
#every bacteria barring 52.5 has substantial growth at 24 hours, so might make the most meaningful comparison to use this timepoint
  
#Another first step would be to get a list of which phages are infected! Set cutoff to more than 50% reduction!
df_infected <- df_reduction_plot %>%
  filter(mean_red_perc > 50) %>%
  filter(bacteria != "LB") %>%
  filter(time_hours == 24) %>%
  group_by(Phage, titre) %>%
  reframe(number_infections = n()) %>%
  mutate(percentage_breadth = 100*(number_infections/25)) %>%
  mutate(Phage = ifelse(Phage == "CPL00272", "Tor", Phage))

#graph doesnt make sense as at 10^2, CPL00950 doesnt inhibit anything. need to add another df where everything is 0 as this is filtered out!

df_cpl00950 <- data.frame(
  Phage = "CPL00950",
  titre = 2,
  number_infections = 0,
  percentage_breadth = 0
)

df_infected <- rbind(df_infected, df_cpl00950)

fig2B <- ggplot(df_infected, aes(x = as.numeric(titre), y = number_infections, colour = Phage)) +
  geom_line(linewidth = 2) +
  geom_point(size = 3) +
  scale_x_continuous("log10(Phage Titre(PFU/mL))") +
  scale_y_continuous("Number of Inhibited Strains (max = 25)", limits = c(0,25), breaks = seq(0, 25, 5)) +
  scale_colour_manual("Phage", values = c("#80b1d3", "#fdb462")) +
  theme_cowplot(font_size = 28)

#figure2C-----------------------------------------------------------------------
#reorganised PBIN
#set wd, load in dictionary, and set in the colour palette I like. Note that Dictionary.csv is not with the raw data to make the code work. Also set seed to 42 for bootstrapping I will do later.
dictionary = read.csv("rawdata/figure2C/Dictionary.csv")

#when I read in each individual dataset, I will be using a function to format it into a specific way, and then join them all together.
#will be defining that function here. 

csv_files <- list.files(
  path = "rawdata/figure2C/GrowthCurveData/", 
  pattern = "\\.csv$", 
  full.names = TRUE
)

#This function reformats each individual CSV to produce a long format that is usable. Remove rows with NAs as some tecan runs may mess up.
reformat <- function(df) {
  colnames(df)[1] <- "Well"
  new_names <- sapply(colnames(df)[-1], function(x) substr(x, 2, nchar(x) - 1))
  colnames(df)[-1] <- new_names
  df_long <- pivot_longer(df, cols = -Well, names_to = "Time_Hours", values_to = "values")
  df_long$Time_Hours <- as.numeric(as.character(df_long$Time_Hours))
  df_long <- mutate(df_long, Time_Hours = round(Time_Hours / 3600 / 0.25) * 0.25)
  df_long <- left_join(df_long, dictionary)
  df_long <- df_long %>% na.omit()
  return(df_long)
}

#This function then applies each csv in the working directory folder to the previous function
#and adds a column that uses the filename to state which strain is in use
reformat_csv <- function(file_name) {
  df <- read.csv(file_name, header = TRUE, row.names = NULL)
  df <- reformat(df)
  file_base_name <- sub("_.*", "", basename(file_name))
  df$bacteria <- file_base_name
  return(df)
}

#use purrr::map_df to apply the reformat_csv function to each CSV file and bind them together
df_all <- map_df(csv_files, reformat_csv)

#it turns out well H12 is contaminated with phage and H9 will be excluded.
df_all <- df_all %>% filter(Well != "H12") %>% filter(Well != "H9")

#going to need to calculate virulence index for each phage, in each host
df_summary <- df_all %>% group_by(Time_Hours, bacteria, phage) %>% 
  reframe(bacteria = bacteria, phage = phage, Time_Hours = Time_Hours, mean_od = mean(values))

#extract blank dataframe and left_join it
blks_df = df_summary %>% filter(phage == "BLK") %>% group_by(Time_Hours) %>% summarise(mean_blk_od = mean(mean_od))
df_summary = df_summary %>% left_join(blks_df, by = c("Time_Hours")) %>% mutate(norm_od = mean_od - mean_blk_od)

#calculate the AUC of the controls, and join this into df_summary
#will fire an equal number of warnings as bacteria it collapses each value into a single column
df_virulence = df_summary %>% filter(phage == "No Phage") %>% group_by(bacteria, Time_Hours) %>% 
  reframe(bacteria = bacteria, mean_ctrl_od = mean(norm_od)) %>% group_by(bacteria) %>%
  reframe(mean_ctrl_area = (AUC_trap=AUC(x=Time_Hours, y=mean_ctrl_od, method='trapezoid')))
df_summary = df_summary %>% left_join(df_virulence)

#calculate virulence index for each well in df_summary
#will fire twice number of warnings as bacteria it collapses each value into a single column
df_summary = df_summary %>% group_by(bacteria, phage) %>% 
  reframe(meanAUC = AUC(x=Time_Hours, y=norm_od, method='trapezoid'), mean_ctrl_area = mean_ctrl_area) %>%
  unique() %>% reframe(bacteria = bacteria, phage = phage, Vp = (1 - (meanAUC/mean_ctrl_area)))

#now to create the PBIN, I need to work out if each phage infects a bacteria:
#if Vp > 0.5, then the phage infects the bacteria. 
df_summary <- df_summary %>% mutate(Infection = ifelse(Vp > 0.5, 1, 0))

#now create the PBIN!!
PBIN <- df_summary %>% select(bacteria, phage, Infection) %>%
  pivot_wider(names_from = phage, values_from = Infection)

#As I am generally interested, lets do a data comparison to calculate host range % of each phage
#Need to sum up each column of the PBIN and divide by the number of rows
PBIN_summary <- PBIN %>% select(-bacteria)
n_rows <- nrow(PBIN)
sums <- colSums(PBIN_summary) / n_rows
PBIN_summary <- data.frame(phage = colnames(PBIN_summary), Percentage_Infected = sums*100)
PBIN_summary <- PBIN_summary %>% arrange(Percentage_Infected)

#Going to make a preliminary heatmap plot of Pseudomonas phages as rows against the hosts as columns, ranked from top to bottom by the row sum of each phage (i.e the most virulent at the top)
#Got to first get PBIN into a decnt format for this, need to be pivoted longer.
#Will need X axis of bacteria, Y axis of phage, and a third column for if the interaction happens.

#need to extract the order of phage infectivity from the PBIN_summary df
Phage_order <- PBIN_summary %>% filter(phage != "BLK")
Phage_order <- unlist(as.vector(select(Phage_order[order(Phage_order$Percentage_Infected),], -Percentage_Infected)))

bacteria_order <- PBIN %>% mutate(Promiscuity = rowSums(across(where(is.numeric)))) %>% select(bacteria, Promiscuity)
bacteria_order <- c(unlist(as.vector(select(bacteria_order[order(-bacteria_order$Promiscuity),], -Promiscuity))))

exclude_impure_and_sedgewick <- c("CPL00934", "CPL00937", "CPL00951", "CPL00954", "CPL00933", "CPL00935", "CPL00953", "CPL00943", "CPL00936", "CPL00938", "CPL00944", "CPL00941", "CPL00939",  "CPL00940", "CPL00942", "CPL00946", "CPL00949", "CPL00952", "CPL00165")

Heatmap_PBIN <- pivot_longer(PBIN, cols = -bacteria, names_to = "Phage", values_to = "Infection") %>%
  mutate(Infection = factor(Infection, levels = c("1", "0"))) %>% 
  mutate(Phage = factor(Phage, levels = Phage_order)) %>%
  filter(Phage != "T7") %>% 
  mutate(bacteria = factor(bacteria, levels = bacteria_order)) %>%
  filter(!Phage %in% exclude_impure_and_sedgewick) 


Heatmap_plot <- ggplot(Heatmap_PBIN, aes(x = bacteria, y = Phage, fill = Infection)) +
  geom_tile(colour = "white", lwd = 1, linetype = 1) +
  scale_fill_manual(values = c("1" = "#bc80bd", "0" = "#8dd3c7"), labels=c('Infection', 'No Infection')) +
  scale_y_discrete('Phage') +
  scale_x_discrete("Pseudomonas aeruginosa Strain") +
  labs(fill = "Phage Infection") +
  theme_classic(base_size=16) +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_rect(colour = "black")) +
  theme(axis.text.x = element_text(angle = 90))

# making the actual heatmap
if (!requireNamespace("BiocManager", quietly=TRUE))
  install.packages("BiocManager")
library(ComplexHeatmap)     #For making dendrogram based heatmap
library(circlize)           #For making dendrogram based heatmap
library(dendextend)         #For making dendrogram based heatmap

#now to try and incorperate genomes of only single phages
#for now change the WD and add in data about each phage from VIRIDIC i.e. genus and species

exclude_impure_and_sedgewick <- c("CPL00056", "CPL00934", "CPL00937", "CPL00951", "CPL00954", "CPL00933", "CPL00935", "CPL00953", "CPL00943", "CPL00936", "CPL00938", "CPL00944", "CPL00941", "CPL00939",  "CPL00940", "CPL00942", "CPL00946", "CPL00949", "CPL00952", "CPL00165")

VIRIDIC <- read.csv("rawdata/figure2C/Comparison/Raw data about each phage no 56.csv") %>%
  filter(!genome %in% exclude_impure_and_sedgewick) %>%
  remove_rownames %>%
  column_to_rownames(var="genome") %>%
  as.matrix()

VIRIDIC_cluster <- read.csv("rawdata/figure2C/Comparison/Phage_clusters.csv") %>% rename(phage = genome)

#Now will need to leftjoin an inverted PBIN onto the VIRIDIC cluster plot after wrangling PBIN plot to work

VIRIDIC_PBIN = PBIN
VIRIDIC_PBIN <- as.data.frame(t(VIRIDIC_PBIN))
VIRIDIC_PBIN <- rownames_to_column(VIRIDIC_PBIN)
names(VIRIDIC_PBIN) <- VIRIDIC_PBIN %>% slice(1) %>% unlist()
VIRIDIC_PBIN <- VIRIDIC_PBIN %>% slice(-1)
VIRIDIC_PBIN <- VIRIDIC_PBIN %>% filter(bacteria != "No Phage" & bacteria != "P56++") %>%
  rename(phage = bacteria)
VIRIDIC_PBIN <- VIRIDIC_cluster %>% left_join(VIRIDIC_PBIN) %>% na.omit() %>%
  filter(!phage %in% exclude_impure_and_sedgewick) 


VIRIDIC_PBIN <- VIRIDIC_PBIN %>% remove_rownames() %>% column_to_rownames(var = "phage") %>%
  select(-genus_cluster, -species_cluster) %>% data.matrix()

#trying to make dendrogram using matrix of raw viridic
hc_rows <- hclust(dist(VIRIDIC))
row_dendro <- as.dendrogram(hc_rows)
row_dendro = color_branches(row_dendro, k = 4)

colors <- c("1" = "#8dd3c7", "2" = "#bc80bd")
#drawing the heatmap
Panel_A <- Heatmap(VIRIDIC_PBIN, name = "Phage Inhibition", col = colors,
                   column_title = "Pseudomonas aeruginosa strain", row_title = "Phage",
                   column_title_side = "bottom",
                   show_column_dend = FALSE, row_dend_side = "right",
                   row_names_side = "left",
                   cluster_rows = row_dendro, 
                   row_dend_width = unit(10, "cm"), column_km = 3,
                   row_split = 3,
                   row_title_gp = gpar(fontsize = 26),
                   column_title_gp = gpar(fontsize = 26),
                   column_names_gp = grid::gpar(fontsize = 26),
                   row_names_gp = grid::gpar(fontsize = 26),
                   heatmap_legend_param = list(
                     at = c(1, 2),  # Specify the discrete numeric values
                     title_gp = gpar(fontsize = 28),
                     labels_gp = gpar(fontsize = 28),
                     labels = c("1" = "No Inhibition", "2" = "Inhibition")  # Customize the labels
                   ))
fig2C <- grid.grabExpr(draw(Panel_A, heatmap_legend_side = "left"))



fig2A <- plot_grid(fig2A_top, fig2A_bottom, labels = NULL, nrow = 2, label_size = 24)  
lower <- plot_grid(fig2A, fig2B, labels = c("A", "B"), rel_widths = c(1,1), label_size = 24)

figure <- plot_grid(lower, fig2C, labels = c("", "C"), nrow = 2, rel_heights = c(1.2,1.5), label_size = 24)

figure
ggsave("PDF-of-figures/figure2.pdf", plot = figure, dpi = 300)

```

{{< pagebreak >}}

#### Tor is potent, broad-range, stable under relevant buffer conditions, and amenable to nebulisation

Once Tor had been identified as an excellent candidate for phage therapy, we next sought to evaluate its phenotypic and genotypic properties. Transmission electron microscopy of negatively stained Tor samples confirmed *Pbunavirus* morphology (Figure 3A), and annotation of its 66.8 kb genome revealed no genes encoding known bacterial virulence factors, antimicrobial resistance genes, or phage genome integration into the host chromosome (Figure S7). Tor possessed 93.5% intergenomic similarity to *Pseudomonas* phage 14-1 which has routinely been used therapeutically (Figure S8; Pirnay 2024). Tor infection parameters include a latent period of 40-50 minutes and a burst size of 60 phages per infected cell (Figure 3B).

Tor was stable in sLB broth, SM buffer, and 0.9% saline over a five-month period at 4 ^o^C and room temperature (Figure 3C). Furthermore, nebulisation of Tor using an Altera eFlow vibrating mesh nebuliser had minimal impact on phage titres (58% recovery, SD = 12%), with the majority of phage titres found within droplet sizes suitable for delivery to the lower lung (median mass aerodynamic diameter = 2.75 µm (SD = 0.12 µm); geometric standard deviation = 1.96 µm) (Figure 3D-E). Overall, this suggests Tor is a prime candidate for therapy, with a high likelihood of safety, nebulised into a wide droplet size range for delivery across the lower lung, routinely propagates to titres in excess of 10^10^ PFU/mL, and possesses sufficient stability for cost effective storage and application.

{{< pagebreak >}}

```{r fig-biology, fig.asp=1.41, fig.width=24, warning=FALSE, message=FALSE, dpi=300, out.height='14.1in', out.width='10in'}
#| fig-cap: "Tor has favourable phenotypic properties for phage therapy. A) Micrograph image of negatively stained Tor. B) One step growth curve of Tor in P. aeruginosa type strain PAO1. The curve and 95% confidence intervals are calculated using local regression with Locally Estimated Scatterplot Smoothing (LOESS). C) Titre of three independent Tor stocks measured in triplicate over a five-month period in each of three buffers: sLB (circle), SM buffer (triangle), and 0.9% saline (square), at two temperatures: 4 degrees celcius (purple) and room temperature (orange). Error bars represent standard deviation. D+E) Titre of Tor (± standard deviation) after nebulisation collected at defined particle size cutoffs, represented as the mean recovered phage particles in D and the cumulative fraction of total recovered phage particles in E. The black vertical line in D and E represents the median particle size calculated by linear interpolation and the grey shaded area represents standard deviation."
#| fig-dpi: 300

# figure3A ---------------------------------------------------------------------
# TEM image of CPL00272
# use magick to read in the tif produced by bioimaging
tem_image <- magick::image_read("rawdata/figure3A/VG_micrograph.png")
#too big, have to resise
tem_image <- magick::image_scale(tem_image, "1000")
#now make the figure
fig3a <- ggplot() + theme_void() + draw_image(tem_image)


#figure3B-----------------------------------------------------------------------
#One-step growth curve data in PAO1

df <- read_csv("rawdata/figure3B/PAO1_272_onestep.csv")

df_plot <- df %>% 
  filter(test == "test") %>% 
  mutate(time = as.numeric(time)) %>%
  mutate(log_pfu = log10(PFU_per_mL))

#calculate average pre-burst size
mean_pre_burst <- df_plot %>%
  filter(time < 45) %>%
  select(log_pfu) %>%
  as.vector() %>%
  unlist() %>%
  mean()

#calculate average post-burst size
mean_post_burst <- df_plot %>%
  filter(time > 55) %>%
  select(log_pfu) %>%
  as.vector() %>%
  unlist() %>%
  mean()

midpoint <- mean_pre_burst + (mean_post_burst - mean_pre_burst)/2

burst_size <- (10^mean_post_burst)/(10^mean_pre_burst)
# print(burst_size) #print size = 60.31149; 60 phages per infection

fig3b <- ggplot(df_plot, aes(x = time, y = log_pfu)) +
  geom_smooth() +
  geom_point(aes(group=replicate), position = position_jitter(width=0.5)) +
  geom_hline(yintercept = mean_pre_burst, linetype = "dashed", colour = "darkgrey") +
  annotate(geom = "label", x = 90, y = mean_pre_burst, label = "Pre-Burst Titre", size = 20/.pt) +
  geom_hline(yintercept = mean_post_burst, linetype = "dashed", colour = "darkgrey") +
  annotate(geom = "label", x = 90, y = mean_post_burst, label = "Post-Burst Titre", size = 20/.pt) +
  annotate(geom = "label", x = 90, y = midpoint, label = "Burst Size", size = 20/.pt) +
  geom_vline(xintercept = 40, linetype = "dashed", colour = "darkgrey") +
  annotate(geom = "label", x = 25, y = 2.0, label = "Latent Period", size = 20/.pt) +
  geom_segment(aes(x = 80, y = mean_pre_burst, xend = 80, yend = mean_post_burst), arrow = arrow(length = unit(0.02, "npc"))) +
  scale_y_continuous("Log10(Phage Titre (PFU/mL)", limits = c(2,4.7), breaks = seq(0,5,0.5)) +
  scale_x_continuous("Time (mins)", breaks = seq(0, 80, 10), limits = c(8,100)) +
  theme_cowplot(28)


#figure3C-----------------------------------------------------------------------

#stability data
df <- read.csv("rawdata/figure3C/Data.csv") %>%
  mutate(titre1 = log10((PFU1*(10^(DF1)))/0.005)) %>%
  mutate(titre2 = log10((PFU2*(10^(DF2)))/0.005)) %>%
  mutate(titre3 = log10((PFU3*(10^(DF3)))/0.005)) %>%
  select(Phage, Month, Buffer, Temperature, Repeat, titre1, titre2, titre3) %>%
  mutate(Repeat = as.factor(Repeat)) %>%
  mutate(meanPFU = (titre1 + titre2 + titre3)/3) %>%
  group_by(Phage, Buffer, Temperature, Month) %>%
  reframe(mean = mean(meanPFU, na.rm= TRUE), sd = sd(meanPFU, na.rm = TRUE)) %>%
  mutate(Temperature = ifelse(Temperature == "RT", "Room Temperature",
                              ifelse(Temperature == "4", "4 \u2103", NA))) %>%
  mutate(Temperature = factor(Temperature, levels = c("4 \u2103", "Room Temperature"))) %>%
  filter(Phage == 272)

fig3c <- ggplot(df, aes(x = Month, y = mean, shape = Buffer, colour = Temperature)) +
  geom_line(size = 1) +
  geom_point(size = 3) +
  scale_colour_manual(values = c("#bebada", "#fb8072"), labels = c("4 °C", "Room Temperature")) +
  scale_y_continuous("Mean Viable Phage Titre (log10 PFU/mL)", limits = c(6, 10), breaks = seq(6,10,1)) +
  geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd), width = 0.2, size = 1) +
  theme_cowplot(28) +
  theme(strip.text.y = element_blank())

#stats
#null = no difference in P272 titre over the time period compared to baseline of SM and 4 degrees
#alternative = difference in phage titre related to time, buffer, or temperature
#threshold alpha = 0.05
#initial test = mixed effects linear regression

#stats
#the value I am interested in is PFU - so itll start titre ~

#decide on fixed effects that I am interested in:
#month, buffer, temperature

#then decide on random effects which are representative of individuals
#repeat (combine into a single term called interaction as repeats were indepenant
#i.e. rpt 1 for 272_lb_RT was not the same sample as 272_sal_4degrees).

# library(nlme)
# df_stats <- read.csv("rawdata/figure4A/Data.csv") %>%
#   filter(Phage != 950) %>%
#   mutate(titre1 = log10((PFU1*(10^(DF1)))/0.005)) %>%
#   mutate(titre2 = log10((PFU2*(10^(DF2)))/0.005)) %>%
#   mutate(titre3 = log10((PFU3*(10^(DF3)))/0.005)) %>%
#   select(Phage, Month, Buffer, Temperature, Repeat, titre1, titre2, titre3) %>%
#   mutate(Repeat = as.factor(Repeat)) %>%
#   mutate(meanPFU = (titre1 + titre2 + titre3)/3) %>%
#   mutate(Interaction = paste(Phage, "_", Repeat)) %>%
#   select(Interaction, Month, Temperature, Buffer, meanPFU) %>%
#   drop_na(meanPFU) %>%
#   mutate(Buffer = factor(Buffer, levels = c("SM", "Saline", "LB"))) %>%
#   mutate(Temperature = factor(Temperature, levels = c("4", "RT")))
# 
# 
# #What this is saying is that I predict meanPFU to be impacted by month and temp,
# 
# model <- lme(meanPFU ~ Month + Temperature * Buffer,
#              random = ~ 1 | Interaction,
#              data = df_stats)
# summary(model)
# plot(model)

#figure3D-----------------------------------------------------------------------
df <- read.csv("rawdata/figure3D+E/CF_Data_for_R.csv") %>%
  na.omit() %>%
  select(-Sample) %>%
  group_by(Particle.Diameter) %>%
  reframe(mean_log_pfu = mean(log10(Number.PFU)), sd_pfu = sd(log10(Number.PFU)), 
          mean_cum_pfu = mean(cumulative.PFU.recovered), sd_cum_pfu = sd(cumulative.PFU.recovered)) %>%
  filter(Particle.Diameter != "MOC") %>%
  mutate(Particle.Diameter = as.numeric(Particle.Diameter))

fig3d <- ggplot(df, aes(x = Particle.Diameter, y = mean_log_pfu)) +
  geom_point(size = 2) +
  geom_line() +
  geom_errorbar(aes(ymin = mean_log_pfu - sd_pfu, ymax = mean_log_pfu + sd_pfu), width = 0.2) +
  scale_y_continuous("Mean Recovered PFU (log10)", limits = c(2,10), breaks = seq(3,9,1)) +
  scale_x_continuous("Aerosolised Particle Size (μm)", breaks = c(0.98,1.36,2.08,3.3,5.39,8.61,14.1)) +
  geom_rect(aes(xmin = 2.76-0.12, xmax = 2.76+0.12, ymin = 3, ymax = 9),
            colour = "lightgrey", alpha = 0.05) +
  annotate("segment", x = 2.75, xend = 2.75, y = 3, yend = 9, linewidth = 1) +
  geom_segment(x = 0.98, y = 9.1, xend = 5, yend = 9.1, 
               colour = "black", 
               arrow = arrow(ends = 'both', length = unit(0.03, "npc")), 
               linewidth = 1) +
  geom_segment(x = 5.1, y = 9.1, xend = 14.1, yend = 9.1, 
               colour = "black", 
               arrow = arrow(ends = 'both', length = unit(0.03, "npc")), 
               linewidth = 1) +
  geom_text(x = 3, y = 9.5, label = "Respirable Fraction", size = 9) +
  geom_text(x = 9.1, y = 9.5, label = "Conducting/Upper Airway Fraction", size = 9) +
  theme_cowplot(28) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

fig3e <- ggplot(df, aes(x = Particle.Diameter, y = mean_cum_pfu)) +
  geom_point(size = 2) + 
  geom_line() +
  geom_errorbar(aes(ymin = mean_cum_pfu - sd_cum_pfu, ymax = mean_cum_pfu + sd_cum_pfu), width = 0.2) +
  scale_y_continuous("Mean Fraction of Total PFU Recovered (Cumulative)", limits = c(-0.1,1.2), breaks = seq(0,1,0.1)) +
  scale_x_continuous("Aerosolised Particle Size (μm)", breaks = c(0.98,1.36,2.08,3.3,5.39,8.61,14.1)) +
  geom_rect(aes(xmin = 2.75-0.12, xmax = 2.75+0.12, ymin = 0, ymax = 1),
            colour = "lightgrey", alpha = 0.05) +
  annotate("segment", x = 2.76, xend = 2.76, y = 0, yend = 1, linewidth = 1) +
  geom_segment(x = 0.98, y = 1.1, xend = 5, yend = 1.1, 
               colour = "black", 
               arrow = arrow(ends = 'both', length = unit(0.03, "npc")), 
               linewidth = 1) +
  geom_segment(x = 5.1, y = 1.1, xend = 14.1, yend = 1.1, 
               colour = "black", 
               arrow = arrow(ends = 'both', length = unit(0.03, "npc")), 
               linewidth = 1) +
  geom_text(x = 3, y = 1.15, label = "Respirable Fraction", size = 9) +
  geom_text(x = 9.1, y = 1.15, label = "Conducting/Upper Airway Fraction", size = 9) +
  theme_cowplot(28) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))


#plot them all
top <- plot_grid(fig3a, fig3b, labels = c("A", "B"), nrow = 1, rel_widths = c(1,2), rel_heights = c(1,1), label_size = 24)

bottom <- plot_grid(fig3d, fig3e, labels = c("D", "E"), nrow = 1, rel_widths = c(1,1), label_size = 24)

figure <- plot_grid(top, fig3c, bottom, nrow = 3, labels = c("", "C"), rel_heights = c(0.7,1,1), label_size = 24)

figure
ggsave("PDF-of-figures/figure3.pdf", plot = figure, dpi = 300)
```

{{< pagebreak >}}

#### Tor demonstrates synergy with antibiotics and other clinically relevant phages

As phages are most likely to be administered as an adjunct to antibiotic therapy (Uyttebroek 2022, Pirnay 2024), the interactions of Tor with meropenem and colistin, two widely used anti-*Pseudomonal* antibiotics, were investigated. Across the tested concentrations, the presence of any concentration of meropenem was sufficient to prevent emergence of Tor-resistant bacteria over the 48-hour period at titres of 10^5^ or 10^6^ (MOIs of \~1 and \~10, respectively). Compared with bacterial growth at up to 1 µg/mL in antibiotic only controls (Figures 4A-B), this suggests synergy that could enhance therapy. There was no evidence for any positive or negative interaction between Tor and colistin, indicating that adjunctive use of both would not impede therapy (Figure

To investigate whether Tor could provide benefits over currently relevant *P. aeruginosa* phages, Tor's host range was compared to the widely used phage PNM (Pirnay 2024), against the bacterial panel used in the original host-range screen. Phage PNM inhibited the growth of one more strain than Tor across the 25-phage panel (Figure 5A), although the phages varied in their efficacy of bacterial control across the strain. PAO1, RDE1, and PIC7.1 were more strongly inhibited by Tor, and PIC15.1, PIB15.1, and RDE4 were more strongly inhibited by PNM (Figure S9). These properties translated into improved cocktail suppression when Tor and PNM were combined (Figure S10), with limited antagonism observed between the phages (Figure S11).

{{< pandefinedgebreak >}}

```{r fig-fig4features, fig.asp=1.41, fig.width=16, warning=FALSE, message=FALSE, dpi=300, out.height='14.1in', out.width='10in'}
#| fig-cap: "Tor's efficacy is enhanced by combining it with antibiotics. A-D) Synograms of Tor in the presence of varying concentrations of meropenem in A+B or colistin in C+D and strain PAO1. Values are recorded as either endpoint reduction percentage in panels A and C or fractional reduction in area under the curve compared to untreated wells (virulence) in B and D as the mean of five biological replicates."
#| fig-dpi: 300

#figure4--------------------------------------------------------------------------
#PAS data

#read in the synogram data
dictionary <- read.csv("rawdata/figure4/Dictionaries for synograms/Dictionary.csv")
plate_contents <- read.csv("rawdata/figure4/Dictionaries for synograms/PLateContents.csv")
poly_dict <- read.csv("rawdata/figure4/Dictionaries for synograms/PolyamineDictionary.csv")

#This function will reformat an individual .CSV into the correct format by modifying the columns, reformatting the time
#adding in the dictionary that tells us what is in each well, and removes NAs for dodgy values
reformat <- function(df_test) {
  df_test <- df_test %>% select(-X)
  df_test[1,1] <- "Well"
  names(df_test) <- lapply(df_test[1, ], as.character)
  df_test <- df_test[-1,] 
  df_test <- pivot_longer(df_test, cols = -Well, names_to = "Time_Hours", values_to = "values") %>%
    filter(Time_Hours != "Time")
  df_test$Time_seconds <- sapply(df_test$Time_Hours, function(x) {
    # Extract hours and minutes using regular expressions
    hours <- as.numeric(sub(".*?(\\d+) h.*", "\\1", x))  # Extract hours
    minutes <- as.numeric(sub(".*?(\\d+) min.*", "\\1", x))  # Extract minutes
    
    # # Replace NA with 0 where no match found
    hours[is.na(hours)] <- 0
    minutes[is.na(minutes)] <- 0
    
    # Convert to seconds
    total_seconds <- (hours * 3600) + (minutes * 60)
    return(total_seconds)
  })
  df_test <- left_join(df_test, dictionary)
  df_test <- df_test %>% na.omit()
}

#This function now will take each of the filenames in the folder, and for each it'll take in
#the file name, run the reformatting function above, add in a column with the plate name
#(taken from the file name) 

produce_df <- function(file_name){
  df <- read.csv(file_name, header = TRUE, row.names = NULL)
  df <- reformat(df)
  file_base_name <- sub("\\.\\w+$", "", basename(file_name))
  df$plate <- file_base_name
  return(df)
}

#get the list of all the csv files in the folder
csv_files <- list.files(
  path = "rawdata/figure4/Synograms and polyamines 9.12.24", 
  pattern = "\\.CSV$", 
  full.names = TRUE
)

#use purrr::map_df to apply the produce_df function to each CSV file and bind them together
df_all <- map(csv_files, produce_df)

#df in list form, this isnt ideal, need a concatenated one
df_list <- function(df_list){
  final_df <- Reduce(full_join, df_list)
  return(final_df)
}
df_all <- df_list(df_all)

#now got to left join what is in what plate. As only doing synograms, 
#will remove plates 16-18

df_all <- left_join(df_all, plate_contents)
df_polyamine <- filter(df_all, Test == "LB"|Test == "Spermine"|Test == "Spermidine")
df_all <- filter(df_all, Test != "LB"&Test != "Spermine"&Test != "Spermidine")

#now to fix some mistakes that I made in the experiment
#1 - plate 6 had cols 9 and 10 wrong way - so wrong abx conc

df_all <- df_all %>%
  mutate(Antibiotic = ifelse(plate == "Plate6" & Column == 9, 32, 
                             ifelse(plate == "Plate6" & Column == 10, 16, Antibiotic)))

#2 - plates 11 and 12 got 5 and 6 wrong way, so wrong abx conc

df_all <- df_all %>%
  mutate(Antibiotic = ifelse(plate == "Plate11" & Column == 5, 2, 
                             ifelse(plate == "Plate11" & Column == 6, 1, Antibiotic)))
df_all <- df_all %>%
  mutate(Antibiotic = ifelse(plate == "Plate12" & Column == 5, 2, 
                             ifelse(plate == "Plate12" & Column == 6, 1, Antibiotic)))

#Now the fun bit of plotting!!!
#Apparently values are characters
df_all <- df_all %>%
  mutate(values = as.numeric(values))

#Make a blank dataframe
df_blks <- df_all %>%
  filter(Content == "BLK") %>%
  rename("blk_od" = "values") %>%
  select(Time_Hours, plate, blk_od) %>%
  group_by(Time_Hours, plate) %>%
  reframe(mean_blk_od = mean(blk_od))

#First to normalise each datapoint
df_all <- df_all %>%
  left_join(df_blks) %>%
  mutate(norm_od = values - mean_blk_od)

#checkplot

# ggplot(df_all, aes(x = Time_seconds, y = norm_od, colour = Test, group = plate)) +
#   geom_line() +
#   facet_grid(Test + Row ~ Column ) +
#   theme_cowplot(12)

#looks like a couple of the tobramycin controls are dodgy

# ggplot(filter(df_all, Test == "Tobramycin"), aes(x = Time_seconds, y = values, colour = plate, group = plate)) +
#   geom_line() +
#   facet_grid(plate + Row ~ Column ) +
#   theme_cowplot(12)

#from this it seems there was a couple wells ins plates 12 and 13 with what will have been a single cell in the controls of a couple of plates. I think its probably safer just to remove it. 

#so remove it, and remake the plots.
df_all <- df_all %>%
  filter(Test != "Tobramycin")

#now we need to calculate for each of colistin and meropenem an endpoint reduction.
df_mero_end_controls <- df_all %>%
  filter(Test == "Meropenem") %>%
  filter(Well == "B02") %>%
  filter(Time_seconds == 167400) %>%
  select(plate, norm_od) %>%
  rename(control_endpoint = norm_od)

#make the reduction one
df_mero_end <- df_all %>%
  filter(Test == "Meropenem") %>%
  left_join(df_mero_end_controls) %>%
  filter(Time_seconds == 167400) %>%
  mutate(red_perc = 100*((control_endpoint-norm_od)/control_endpoint)) %>%
  group_by(Well, Antibiotic, Phage) %>%
  reframe(mean_red_perc = mean(red_perc))

#now plot it.
fig4a <- ggplot(df_mero_end, aes(x = factor(log10(Phage)), y = factor(Antibiotic), fill = mean_red_perc)) +
  geom_tile() +
  scale_x_discrete("Log 10 Phage Titre (PFU/mL)", labels = c(0,2,3,4,5,6)) +
  scale_y_discrete("Antibiotic Concentration \u03BCg/mL") +
  scale_fill_gradient2("Mean\nReduction\nPercentage", high = "white", low = "darkorange", midpoint = 50) +
  theme_cowplot(28)

#now we need to calculate AUC of mero controls.
df_mero_auc_controls <- df_all %>%
  filter(Test == "Meropenem") %>%
  filter(Well == "B02") %>%
  group_by(plate) %>%
  mutate(control_auc = DescTools::AUC(Time_seconds, norm_od, method = "trapezoid")) %>%
  select(plate, control_auc) %>%
  unique()

#make the virulence one
df_mero_auc <- df_all %>%
  filter(Test == "Meropenem") %>%
  group_by(Well, plate, Antibiotic, Phage) %>%
  reframe(auc = DescTools::AUC(Time_seconds, norm_od, method = "trapezoid")) %>%
  left_join(df_mero_auc_controls) %>%
  mutate(Vp = 1 - (auc/control_auc)) %>%
  group_by(Antibiotic, Phage) %>%
  reframe(mean_vp = mean(Vp))

fig4b <- ggplot(df_mero_auc, aes(x = factor(log10(Phage)), y = factor(Antibiotic), fill = mean_vp)) +
  geom_tile() +
  scale_x_discrete("Log 10 Phage Titre (PFU/mL)", labels = c(0,2,3,4,5,6)) +
  scale_y_discrete("Antibiotic Concentration \u03BCg/mL") +
  scale_fill_gradient2("Mean\nPhage\nVirulence", high = "white", low = "darkgreen", midpoint = 0.8) +
  theme_cowplot(28)

#now to make the colistin versions
df_col_end_controls <- df_all %>%
  filter(Test == "Colistin") %>%
  filter(Well == "B02") %>%
  filter(Time_seconds == 167400) %>%
  select(plate, norm_od) %>%
  rename(control_endpoint = norm_od)

#make the reduction one
df_col_end <- df_all %>%
  filter(Test == "Colistin") %>%
  left_join(df_col_end_controls) %>%
  filter(Time_seconds == 167400) %>%
  mutate(red_perc = 100*((control_endpoint-norm_od)/control_endpoint)) %>%
  group_by(Well, Antibiotic, Phage) %>%
  reframe(mean_red_perc = mean(red_perc))

#now plot it.
fig4c <- ggplot(df_col_end, aes(x = factor(log10(Phage)), y = factor(Antibiotic), fill = mean_red_perc)) +
  geom_tile() +
  scale_x_discrete("Log 10 Phage Titre (PFU/mL)", labels = c(0,2,3,4,5,6)) +
  scale_y_discrete("Antibiotic Concentration \u03BCg/mL") +
  scale_fill_gradient2("Mean\nReduction\nPercentage", high = "white", low = "darkorange", midpoint = 50) +
  theme_cowplot(28)

#now we need to calculate AUC of colistin controls.
df_col_auc_controls <- df_all %>%
  filter(Test == "Colistin") %>%
  filter(Well == "B02") %>%
  group_by(plate) %>%
  mutate(control_auc = DescTools::AUC(Time_seconds, norm_od, method = "trapezoid")) %>%
  select(plate, control_auc) %>%
  unique()

#make the virulence one
df_col_auc <- df_all %>%
  filter(Test == "Colistin") %>%
  group_by(Well, plate, Antibiotic, Phage) %>%
  reframe(auc = DescTools::AUC(Time_seconds, norm_od, method = "trapezoid")) %>%
  left_join(df_col_auc_controls) %>%
  mutate(Vp = 1 - (auc/control_auc)) %>%
  group_by(Antibiotic, Phage) %>%
  reframe(mean_vp = mean(Vp))

fig4d <- ggplot(df_col_auc, aes(x = factor(log10(Phage)), y = factor(Antibiotic), fill = mean_vp)) +
  geom_tile() +
  scale_x_discrete("Log 10 Phage Titre (PFU/mL)", labels = c(0,2,3,4,5,6)) +
  scale_y_discrete("Antibiotic Concentration \u03BCg/mL") +
  scale_fill_gradient2("Mean\nPhage\nVirulence", high = "white", low = "darkgreen", midpoint = 0.8) +
  theme_cowplot(28)


#to plotting. Stability one can go aloing top, then the two nebs side by side, and finally the four synograms.

figure <- plot_grid(fig4a, fig4b, fig4c, fig4d, nrow = 2, ncol = 2, labels = "AUTO", label_size = 20)

figure
ggsave("PDF-of-figures/figure4.pdf", plot = figure, dpi = 300)
```

{{< pagebreak >}}

#### Tor host range can be expanded through co-evolution driving tail modifications

Tor did not initially appear to inhibit growth of strain DFI10.1, alhough we observed occasional, single plaques on bacterial lawns when Tor was spotted at high titre, suggesting the presence of rare mutant phages with the ability to infect this strain. These plaques were picked and propagated on DFI10.1 to produce TorV1.1, which improved plaque appearance from hazy, indicative of weak infection or lysis from without, to a full zone of lysis, similar to that of Tor on PAO1 (Figure 5B). This improved infection ability also translated into improved inhibition of bacterial growth across a range of multiplicities of infection in liquid culture (Figure 5C).

Sequencing of the TorV1.1 genome revealed a single non-synonymous substitution in gp48, an annotated tail protein (N926Y), that was present in 100% of the mapped reads (Tor: PZ447396; TorV1.1: PZ447397). To investigate how the observed substitution may impact tail structures, we used AlphaFold3 to predict the structure of the trimeric Tor and TorV1.1 tail fiber proteins. The structures revealed that the amino acid change occurred at the C-terminal tip, which critically is involved in receptor binding with the host through interactions with carbohydrate domains in the related *Pbunavirus* E217 (Figure 6; Li 2023). Therefore, the difference in host-range is likely explained by a small but meaningfully altered phage-host binding interaction, altering which *P. aeruginosa* strains are susceptible to infection.

Tor is therefore an adaptable phage that spontaneously produces rare variants with altered host ranges, providing clincial flexibility through selection of appropriate plaques. Screening of these rare mutants across future clinical isolates could be used to identify phage variants for inclusion in broad-spectrum phage cocktails or panels.

{{< pagebreak >}}

```{r fig-fig5future, fig.asp=0.72, fig.width=16, warning=FALSE, message=FALSE, dpi=300, out.height='8.64in', out.width='12in'}
#| fig-cap: "Enhancing the efficacy of Tor through phage-phage interactions and host-range expansion. A) Breadth of phages Tor, PNM and a 1:1 cocktail as measured over a 48-hour period. A strain was deemed susceptible if the area under the curve for a bacterial strain was reduced by over 50%. B) Plaque assays of Pseudomonas phage Tor and its variant on Pseudomonas aeruginosa strains PAO1 and DFI10.1. C) Growth curves of phage Tor and its variant on Pseudomonas aeruginosa strain DFI10.1 facetted by multiplicity of infection."
#| fig-dpi: 300

#figure 5a - PNM and Tor

#read in the synogram data
row_content <- read.csv("rawdata/figure5A/row_data.csv")
col_content <- read.csv("rawdata/figure5A/column_data.csv")

#This function will reformat an individual .CSV into the correct format by modifying the columns, reformatting the time
#adding in the dictionary that tells us what is in each well, and removes NAs for dodgy values
reformat <- function(df_test) {
  df_test <- df_test %>% select(-Well)
  df_test[1,1] <- "Well"
  #names(df_test) <- lapply(df_test[1, ], as.character)
  df_test <- df_test[-1,] 
  df_test <- rename(df_test, "Time_Hours" = "X")
  
  df_test <- pivot_longer(df_test, cols = -Time_Hours, names_to = "Well", values_to = "values") %>%
    filter(Time_Hours != "Time")
  
  df_test <- df_test %>%
    mutate(
      hours = as.numeric(str_extract(Time_Hours, "\\d+(?=\\s*h)")),
      minutes = as.numeric(str_extract(Time_Hours, "\\d+(?=\\s*min)")),
      hours = replace_na(hours, 0),
      minutes = replace_na(minutes, 0),
      Time_seconds = (hours * 3600) + (minutes * 60)
    ) %>%
    select(-hours, -minutes)  # Remove temporary columns
  
  df_test <- df_test %>%
    mutate(
      Row = str_extract(Well, "^[A-Z]"),
      Column = as.numeric(str_extract(Well, "\\d+$"))
    )
}


#This function now will take each of the filenames in the folder, and for each it'll take in
#the file name, run the reformatting function above, add in a column with the plate name
#(taken from the file name) 

produce_df <- function(file_name){
  df <- read.csv(file_name, header = TRUE, row.names = NULL)
  df <- reformat(df)
  file_base_name <- sub("\\.\\w+$", "", basename(file_name))
  df$plate <- file_base_name
  return(df)
}

#get the list of all the csv files in the folder
csv_files <- list.files(
  path = "rawdata/figure5A/", 
  pattern = "\\.CSV$", 
  full.names = TRUE
)

#use purrr::map_df to apply the produce_df function to each CSV file and bind them together
df_all <- map(csv_files, produce_df)

#df in list form, this isnt ideal, need a concatenated one
df_list <- function(df_list){
  final_df <- Reduce(full_join, df_list)
  return(final_df)
}
df_all <- df_list(df_all)

df_all <- left_join(df_all, row_content) %>% left_join(col_content) %>% na.omit() %>%
  mutate(Row = factor(Row, levels = c("A","B","C","D","E","F","G"))) %>%
  mutate(Column = factor(Column, levels = seq(2,10,1)))

temporary_plot <- ggplot(df_all, aes(x = Time_seconds, y = values, group = Well)) +
  geom_line() +
  facet_grid(plate+Row~Column)

#data handling to make the rest of it make sense
df_all_analysis <- df_all %>%
  mutate(values = as.numeric(values)) %>%
  filter(Time_Hours != "")

#extract blanks
df_blanks <- df_all_analysis %>%
  filter(bacteria == "None") %>%
  filter(phage == "No Phage") %>%
  group_by(Time_seconds, plate) %>%
  mutate(mean_blk_od = mean(values)) %>%
  select(Time_seconds, plate, mean_blk_od)

#normalise od600 and set time to be in hours
df_all_analysis <- df_all_analysis %>%
  left_join(df_blanks) %>%
  mutate(norm_od = values - mean_blk_od) %>%
  filter(bacteria != "None") %>%
  mutate(Time_Hours = Time_seconds/(60*60)) %>%
  mutate(phage = factor(phage, levels = c("No Phage", "Tor", "PNM", "Cocktail"))) 

#plot all the growth curves
# gc_plot <- ggplot(df_all_analysis, aes(x = Time_Hours, y = abs(norm_od), group = Well, colour = phage)) +
#   geom_line() +
#   facet_wrap(~bacteria, nrow = 5) +
#   scale_x_continuous("Time (Hours)", breaks = seq(0,48, 12)) +
#   scale_y_continuous("Normalised OD600") +
#   scale_colour_manual("Phage", values = okabe) +
#   theme_cowplot(20)

#calculate AUC for all control growth curves
control_auc_df <- df_all_analysis %>%
  filter(phage == "No Phage") %>%
  group_by(phage, bacteria, Well) %>%
  mutate(control_auc = DescTools::AUC(Time_Hours, norm_od, method = "trapezoid")) %>%
  group_by(bacteria) %>%
  reframe(mean_cntrl_auc = mean(control_auc)) %>%
  select(bacteria, mean_cntrl_auc)

#calculate virulence for each 
virulence_df <- df_all_analysis %>%
  group_by(phage, bacteria, Well) %>%
  mutate(auc = DescTools::AUC(Time_Hours, norm_od, method = "trapezoid")) %>%
  select(Well, phage, bacteria, auc) %>%
  unique() %>%
  left_join(control_auc_df) %>%
  mutate(vp = 1-(auc/mean_cntrl_auc)) %>%
  mutate(phage = factor(phage, levels = c("No Phage", "Tor", "PNM", "Cocktail")))

virulence_means <- virulence_df %>%
  group_by(phage, bacteria) %>%
  reframe(mean_vp = mean(vp), stdev = sd(vp))
  
virulence_means_plotting <- virulence_means %>%
  filter(phage != "No Phage") %>%
  mutate(does_infect = ifelse(mean_vp > 0.499, "Inhibition", "No Inhibition")) %>%
  mutate(does_infect = factor(does_infect, levels = c("Inhibition", "No Inhibition"))) %>%
  mutate(mean_vp = ifelse(mean_vp < 0, 0, mean_vp))

# host_range_pnm_272 <- ggplot(virulence_means_plotting, aes(x = phage, y = bacteria, fill = mean_vp)) +
#   geom_tile() +
#   scale_fill_gradient2("Mean\nVirulence", high = "#a002a2", mid = "#f7d0f5", low = "white", midpoint = 0.6) +
#   scale_y_discrete('Pseudomonas aeruginosa Strain') +
#   scale_x_discrete("Phage") +
#   labs(fill = "Inhibition") +
#   theme_classic(base_size=20) +
#   theme(panel.grid.major = element_blank(), 
#         panel.grid.minor = element_blank(),
#         panel.background = element_rect(colour = "black")) +
#   theme(axis.text.x = element_text(angle = 90))
# 
# only_infect <- ggplot(filter(virulence_means_plotting, does_infect == "Inhibition"), aes(x = phage, y = mean_vp)) +
#   geom_boxplot() +
#   geom_point(position = position_dodge2(width = 0.4)) +
#   scale_y_continuous("Mean Virulence Index", limits = c(0,1.5), breaks = seq(0,1.5,0.2)) +
#   scale_x_discrete("Phage") +
#   scale_colour_manual("Inhibtion\nover 48h", values = okabe) +
#   theme_cowplot(20)

quick_test_272 <- virulence_means_plotting %>%
  filter(does_infect == "Inhibition") %>%
  select(bacteria, phage, mean_vp) %>% 
  pivot_wider(names_from = phage, values_from = mean_vp) %>%
  na.omit()

# t.test(quick_test_272$Tor, quick_test_272$PNM, paired = TRUE) #no sig difference
# t.test(quick_test_272$Tor, quick_test_272$Cocktail, paired = TRUE) #no sig difference
# t.test(quick_test_272$PNM, quick_test_272$Cocktail, paired = TRUE) #no sig difference

count_data <- virulence_means_plotting %>%
  filter(does_infect == "Inhibition") %>%
  group_by(phage) %>%
  reframe(infect_number = n()) %>%
  mutate(phage = factor(phage, levels = c("Tor", "PNM", "Cocktail")))

fig5a <- ggplot(count_data, aes(x = phage, y = infect_number)) +
  geom_col(fill = "#e28dbe") +
  scale_x_discrete("Phage") +
  scale_y_continuous("Number Strains\nInhibited (Vp > 0.5)", limits = c(0,24), breaks = seq(0, 24, 3)) +
  theme_cowplot(28)

#figure 5B - 272 variant plaques------------------------------------------------

#use magick to read in the svg produced in inkscape
plaques <- magick::image_read("rawdata/figure5B/image_for_manscript.png")
#resise the file as it is currently massive
plaques <- magick::image_scale(plaques, "900")
#create an empty ggplot and then draw the synteny image over the top of it
fig5b <- ggplot() + theme_void() + draw_image(plaques)

#--------------------------------------------------------------------------------

#figure 5C - TorV1.1 Growth Curve
variant_gc <- read.csv("rawdata/figure5C/26-03-03 09-33-50 Single whole plate od600 24 hour.CSV")
variant_dictionary <- read.csv("rawdata/figure5C/dictionary.csv")

variant_gc_modified <- variant_gc %>%
  slice(-1) %>%
  select(-Well) %>%
  pivot_longer(!X, names_to = "Well", values_to = "od600") %>%
  mutate(
    hours = as.numeric(str_extract(X, "\\d+(?=\\s*h)")),
    minutes = as.numeric(str_extract(X, "\\d+(?=\\s*min)")),
    hours = replace_na(hours, 0),
    minutes = replace_na(minutes, 0),
    Time_seconds = (hours * 3600) + (minutes * 60)
  ) %>%
  select(-hours, -minutes) %>%
  left_join(variant_dictionary) %>%
  filter(content != "protector") %>%
  select(-X) %>%
  mutate(od600 = as.numeric(od600))

variant_gc_blks <- variant_gc_modified %>%
  filter(content == "blk") %>%
  na.omit() %>%
  rename(blk_od = od600) %>%
  group_by(Time_seconds) %>%
  reframe(mean_blk_od = mean(blk_od))

variant_gc_modified <- variant_gc_modified %>%
  left_join(variant_gc_blks) %>%
  mutate(norm_od600 = od600 - mean_blk_od) %>%
  filter(content != "blk")

#when plotted controls looked very blank, happy to exclude

variant_gc_modified <- variant_gc_modified %>%
  filter(content != "tor_only" & content != "torv1.1_only") %>%
  mutate(log_titre = titre/(10^5)) %>%
  filter(str_detect(Well, "^*7") == FALSE)

fig5c <- ggplot(variant_gc_modified, aes(x = (Time_seconds/3600), y = norm_od600, colour = phage, group = Well)) +
  geom_line(linewidth = 1) +
  facet_wrap(~factor(log_titre), ncol = 3) +
  scale_x_continuous("Time (hours)", limits = c(0,24), breaks = seq(0,24,6)) +
  scale_y_continuous("OD600", limits = c(-0.02, 1.4), breaks = seq(0, 1.5, 0.3)) +
  scale_colour_manual("Treatment", values = c("lightblue", "grey"), labels = c("Tor", "Tor V1.1")) +
  theme_cowplot(28)

#---------------------------------------------------------------------------------
bottom <- plot_grid(fig5b,fig5c, nrow = 1, ncol = 2, labels = c("B", "C"), rel_heights = c(1, 0.5), rel_widths = c(0.8,1.1), label_size = 24)

figure <- plot_grid(fig5a, bottom, nrow = 2, labels = c("A", ""), rel_heights = c(1,1), label_size = 24)

figure
ggsave("PDF-of-figures/figure5.pdf", plot = figure, dpi = 300)
```

{{< pagebreak >}}

```{r fig-fig6tails, fig.asp=1.41, fig.width=12, warning=FALSE, message=FALSE, dpi=300, out.height='11.66in', out.width='8.27in'}
#| fig-cap: "Structural predictions of Tor and TorV1.1 tail fibres (gp48). A) The trimeric Tor gp48 structure, shown in side view and enlarged side and 90° rotated views with each chain coloured differently. Asparagine 926 is highlighted in red. The same side view is shown underneath and coloured by per-residue pLDDT score according to the model confidence key, with corresponding ipTM and pTM values shown below each section with the predicted alignment error (PAE). The N-terminal 20 residues of section 1 (residues 1 to 363) are disordered and thus display low confidence. B)	AlphaFold prediction of the trimeric TorV1.1 gp48 structure, shown in side view and enlarged 90° rotated views with each chain coloured differently. Tyrosine 926 is highlighted in red. The same side view is shown underneath and coloured as in panel A, with corresponding ipTM and pTM values shown below each section with the PAE shown. The N-terminal 23 residues of section 1 (residues 1 to 363) are disordered and thus display low confidence."
#| fig-dpi: 300

#use magick to read in the png produced by phold and edited in inkskape
tails_all <- magick::image_read("rawdata/figure6/Tor_tail_fibre_predictions_2026_revision2_JC_VG.png")
#too big again, need to rescale
tails_all <- magick::image_scale(tails_all, "1000")
#now make the figure
figure <- ggplot() + theme_void() + draw_image(tails_all)

figure
ggsave("PDF-of-figures/figure6.pdf", plot = figure, dpi = 300)
```

## Discussion

As phage biobanks increase in size, the number of pairwise comparisons to experimentally evaluate phage host range increases quadratically, raising issues of scalability. With phage biobanks often now possessing over 1000 phage stocks (resch 2024), current strategies for designing broad-spectrum phage cocktails or panels often include phage down-selection based on informatically informed features (i.e. intergenomic similarity) or limited phenotypic data (i.e. receptor binding proteins, cross-resistance). Whilst designing broad-spectrum phage products using these features has been successful previously (Weiner 2025; Matrinez-Soto 2024, Kim 2024), we postulated that such an approach could arbitrarily screen out highly potent phages belonging to the same down-selection group. Here, we demonstrate that empirical evidence is currently required to select the best phage candidates, particularly between closely related phages.

A pairwise screen of 37 phages and 25 predominantly respiratory *P. aeruginosa* isolates identified a novel and potent phage (Tor) with optimal characteristics for phage therapy, which may have been missed using predictive methods. Its superior breadth over the *P. aeruginosa* panel, identified its high potential for treating patients with respiratory infections. Its genome showed no evidence of features of concern for phage therapy such as the presence of genes encoding lysogeny, virulence factors, or antibiotic resistance genes (cook and hynes 2025). Furthermore, its intergenomic similarity to *Pseudomonas* phage 14-1, a safe, clinically used phage (Pirnay 2024) present a high chance this safety profile extends to Tor, a potential subject for future clinical trials. Phenotypic characterisation of Tor confirmed it did not suffer from previously identified constraints for viable phage products, namely phage instability (Jault 2021), antagonism with standard-of-care antibiotics (Kever, zuo), or incompatibility with respiratory application methods, in this case nebulisation (leung, thompson).

To increase the breadth and efficacy of phage products, it is unlikely phages will be used in isolation. Combining different phages into one broad-spectrum product or a panel for susceptibility testing prior to treatment is central to many phage therapy strategies (magistral, yale, georgia). Supporting this idea, a combination of Tor and the widely used type IV pilus phage PNM (Merabishvilli 2009; Pirnay 2024) improved the spectrum of activity to greater than either phage alone, with some phages better controlling particular strains. Furthermore, Tor's efficacy was improved when using antibiotics and addition of other phages (see also Tong 2026). Host-range expansion via co-evolution with target pathogens was also observed through selection of tail fibre mutants with modified host specificity domains. Therefore, Tor can serve as a strong base to further develop broad-spectrum phage cocktails, onto which other phages, antibiotics, or alternative strategies can be utilised to produce a final, optimised phage product.

In summary, broad-spectrum phage therapy is a strategy with large potential, be they delivered as one product or a panel for screening patient isolates on. The associated issues regarding informatic- or phenotypic-based phage selection were shown here, suggesting that direct evidence of phage breadth is currently required to support inclusion. In the future, machine learning approaches may eliminate this need, although empirical, curated datasets of phage-bacteria interactions determined by defined protocols are required to train these (malajczuk 2026). Currently, to identify more useful phages such as Tor from collections, we recommend scaling automated approaches (i.e. Penke 2026) to generate detailed phage-bacteria interaction networks that can be used to design future panels and cocktails.

## Methods

More detailed methods can be found in Supplementary Text ST1.

#### Bacterial culture, phage, antibiotics, and DNA extraction

Bacteria were collected as clinical isolates from the Royal Devon University Hospital NHS Foundation Trust, or type strains purchased from the NCTC, with PAO1 being provided by the University of Exeter Bacterial Pathogenicity Research Group. The phages used in this study were isolated using the standard Citizen Phage Library methods (Fletcher 2024, Olsen 2020) from environmental and wastewater sources. Antibiotics were prepared in 10 mL volume of autoclaved Milli-Q water at 100 × final concentration, filter sterilised, and frozen in 1 mL aliquots for subsequent use. Phage and bacterial DNA were extracted by using the Norgen phage DNA extraction kit (Norgen Biotek, Thorold, ON, Canada) and Nanobind HMW DNA extraction kit (PacBio, Menlo Park, CA, USA), respectively.

#### Phage host range screening

Planktonic Killing Assays (PKAs) were performed to generate phage host range data. As above, bacteria were grown to exponential phase, diluted in fresh sLB to OD~600 nm~ = 0.1, and 5 μL of this was added to 90 μL sLB per well. 5 μL of each well within a phage stock plate, including T7 negative controls, was added to create a final bacterial start OD~600 nm~ of approximately 0.005. The plate was placed in a Sunrise microplate reader (Tecan, Mӓnnedorf, Switzerland) set to 30 ^o^C and optical density at 600 nm was recorded every 15 minutes for 24 hours as a proxy for bacterial growth. For phage-infected wells, local virulence, V~*i*~ (Storms *et al.*, 2020), was calculated over the entire 24-hour period and successful phage inhibition was assumed if V~*i*~ \> 0.5. Given the higher throughput nature of the assay, the number of tested interactions and the stringent V~*i*~ cutoff, this experiment was performed in single replicates per host-virus pair. This process was repeated for 25 *P. aeruginosa* strains in our collection.

All other host-range determination methods were performed either by PKA in a Clariostar Plus Plate Reader with Stacker (BMG LabTech, Ortenberg, Germany) using endpoint OD~600~ or local virulence as appropriate for liquid culture interactions, or efficiency of plating/phage enumeration approaches to calculate titre of each phage on solid media by spotting serially diluted phage stock on exponential host lawns generated by double agar overlays and plaque enumeration.

#### Phage one-step growth curve, stability, nebulisation, and antibiotic synergy testing

One-step growth curves were generated by adding phage stock at a final MOI of 0.01 to exponential-phase PAO1. This was immediately titred, incubated for 3 minutes at 37 ^o^C to allow phage adsorption, and then diluted 1000-fold, returned to a 37 ^o^C water bath, and samples taken every 10 minutes and titred. One-step assays were performed over three biological replicates and burst size calculated as the mean difference between pre- and post-burst, and latent period calculated as the time until onset of lysis.

Phage stability was tested by creating three phage stocks, buffer exchanging them into either SM buffer, 0.9% saline, or fresh sLB using 15 mL Amicon Ultracentrifugal Filters with 100 kDa molecular weight cutoffs (Millipore, Merck, Darmstadt, Germany) through three full washes. Phages were titred and adjusted to 2x10^8^ PFU/mL, creating three independent phage stocks. Phages were stored in 2 mL aliquots in glass amber bottles at either room temperature or 4 ^o^C and 10 µL of each stock was titred at months 1, 2, and 5 in three technical replicates.

The nebulisation efficacy of Tor was assessed as in Thompson et al (2024). Briefly, phages were adjusted to 1.5x10^9^ PFU/mL in sterile water, and nebulised using an Altera eFlow vibrating mesh nebuliser (PARI Respiratory Equipment, Midlothian, VA, USA) for 7 minutes into a refrigerated Next Generation cascade Impactor (MSP, Minneapolis, MN, USA) using a vacuum rate of 15 L/min. This collected aerosols of different sizes into collection cups that were then titred to calculate the number of phages within each aerosol droplet size fraction. Median mass aerodynamic diameter, geometric standard deviation, and recovery percentage were calculated as before (Thompson et al 2024).

Interaction with antibiotics was assessed via synography (Liu et al 2020) over a 46.5-hour period. Antibiotic concentrations were increased across the length of a 96-well-plate, and phage concentration increases across its width. By comparing endpoint reduction of phage-antibiotic treated wells to untreated bacteria-only controls, as in liu et al, as well as area under the curve in the form of local virulence, a phage's ability to control a bacterial population is assessed as phage and antibiotic concentration changes.

#### Phage tail fibre structure prediction

The Tor tail fibre is composed of the 987 amino acid residue protein gp48, which exists as a homotrimer. The TorV1.1 variant differs by a single amino acid substitution, in which an asparagine residue is replaced by a tyrosine at position 926. To visualise this substitution in the structural context of the tail fibre, we predicted the structures of the Tor and Tor v1.1 gp48 proteins using AlphaFold3 (Abramson 2024).  

Prediction of the full-length gp48 trimer did not yield a reliable model. Therefore, the tail fibre protein was divided into three sections, each of which were predicted independently as trimers. Section 1 comprised residues 1 to 363, section 2 residues 364 to 689, and section 3 residues 690 to 987. The top five outputs for each section were similar, therefore only the most confident model is presented.  The resulting pLDDT scores indicate overall confident predictions on the per-atom scale, particularly at the tip of the tail fibre. The corresponding ipTM and pTM values (a measure of the accuracy of the predicted structure) were also within acceptable ranges. Specifically, pTM scores above 0.5 are considered acceptable, whereas ipTM scores of \<0.6 suggests failed predictions.  The three trimeric sections were subsequently aligned manually in ChimeraX (Meng, 2023) to generate a composite model of the full-length tail fibre extending from the base to the tip.

#### Data representation and statistical analysis

Data analysis was performed in R (v4.1.1) and initial statistical tests were simple linear regressions with a threshold alpha throughout of 0.05. Where applicable for categorical data, the baseline of the regression was untreated controls, and where log values are included, 1e-12 was added to ensure no infinities were included in the linear model. Construction of the manuscript, including scripts and raw data to create initial figures and perform statistical analysis, was produced in Quarto v1.4.554 in RStudio 2.023.06.0 Build 421 and rendered as a .docx file using knitr v1.5.0, available at: *insert github link here*. Details about packages can be found in Supplementary Text ST2.

## References

::: {#refs}
:::

## Supplementary

### Supplementary Text ST1 (Methods)

#### General Bacterial Methods

Bacterial strains originating from the Citizen Phage Library were provided by the Royal Devon and Exeter Hospital, streaked onto 1.5% lysogeny broth (LB) agar, grown overnight at 37 ^o^C, and cryostocks were subsequently made using Pro-Lab Microbank vials (Fisher Scientific, Loughborough, UK). Those originating from the International Strain Panel were provided by Cardiff University and similarly streaked and grown but additionally grown overnight in 12 mL LB broth supplemented with 10 mM MgCl~2~ and 10 mM CaCl~2~ (hereafter referred to as sLB). 1 mL of the resulting suspension was adjusted to contain 25% *w/v* glycerol. All strains were stored at -70 ^o^C and streaked out when required onto 1.5% LB agar and left for either 24 or 48 hours depending on the growth rate of the strain. To generate exponential (OD~600~ = 0.55-0.65) or overnight bacterial cultures whilst avoiding accidental propagation of rare, less fit phenotypes, a sweep of bacterial colonies was taken and grown in sLB broth. 

#### Bacterial DNA extraction, sequencing, assembly, and core genome phylogeny

Bacteria were grown in sLB to an OD~600~ of 1 and high molecular weight (HMW) DNA was extracted using a Nanobind HMW DNA extraction kit (PacBio, Menlo Park, CA, USA) and quantified using the Qubit 4 and associated broad range assay kit (Invitrogen, Waltham, MA, USA). Library prep was performed using the rapid sequencing DNA V14 - barcoding (SQK-RBK114.96; ONT, Oxford, UK) kit and long read sequencing was performed using the R10.4.1 MinION flow cells (ONT, Oxford, UK). Using the median read length and an estimated genome size of 6.5 Mb, if coverage from the runs was less than 80-fold, reads were assembled using Unicycler, if estimated coverage was over 80-fold then the Trycycler (https://github.com/rrwick/Trycycler) pipeline was used. Final genomes were then error-corrected using Medaka-11.3 and annotated using Prokka v1.14.6. Core genome phylogeny on associated gff3 files of strains from this study and the previously published international *Pseudomonas* reference panel (Freschi *et al.*, 2018) was performed using Roary (Page *et al.*, 2015), aligning bacterial genome sequences using PRANK, and core genomes using MAFFT. Maximum likelihood trees were produced using FastTree (Price, Dehal and Arkin, 2010) and visualised in R (v4.4.1) using ggTree from the treeio (Wang *et al.*, 2020) package (v1.29.0). 

#### General phage methods

Phages within the Citizen Phage Library are stored as stocks in glass amber bottles at 4 ^o^C and when needed, propagated from stocks by adding 50 μL phage to 20 mL sLB and 500 μL overnight bacterial host. After overnight growth, bacteria were removed by centrifugation (10,000 × g for 30 minutes at 4 ^o^C) and filter sterilisation through a 0.22 μm PES syringe filter (Sartorius, Goettingen, Germany). Phage titres were determined by double agar overlay: creating a lawn of bacteria by mixing exponential bacteria in a 1:3 ratio with molten top agar (0.65% agar, 10 mM MgCl~2~, 10 mM CaCl~2~) and pouring over pre-warmed bottom agar plates (1% agar, 10 mM MgCl~2~,10 mM CaCl~2~). Once set, 10 μL each of a series of 1:10 serial dilutions of phage suspension were spotted onto the lawn in triplicate, dried, and left overnight at 37 ^o^C prior to enumeration of plaque forming units (PFU).

#### Phage DNA extraction, sequencing, assembly, and analysis

Phages were propagated in 20 mL volumes and filter sterilised as above and treated with 20 μL 10 mg/mL DNase I (Roche, Merck, Darmstadt, Germany) and 10 mg/mL RNase A (Invitrogen, ThermoFisher Scientific, Waltham, MA, USA) mixture for 30 minutes at 37 ^o^C. To this, 20% *w/v* PEG8000 and NaCl to a final concentration of 2.5 M was dissolved and left overnight at 4 ^o^C. The mixtures were centrifuged at 10,000 × g for 15 minutes at 4 ^o^C, the supernatants were poured off and resultant pellets resuspended in 750 μL SM buffer (100 mM NaCl, 50 mM Tris-HCl; 8 mM MgSO~4~·H~2~O). Phage DNA was extracted using the Norgen phage DNA extraction kit (Norgen Biotek, Thorold, ON, Canada) following the manufacturer's instructions, quantified using the Qubit 4 and associated broad range assay kit (Invitrogen, Waltham, MA, USA). DNA libraries were prepared by the Exeter Sequencing Service for short read sequencing on the Illumina NovaSeq to generate 150 bp paired-end reads. The BBTools suite (v38.96; Bushnell, 2014) was used to remove reads associated with the host, optical duplicates, adapters, and artefacts from reads before error correction using Tadpole. Reads were first assembled with Shovill (v1.1.0; <https://github.com/tseemann/shovill>) and subsampled using the `–minlength 10,000 and –mincov 20` arguments to avoid errors associated with over-sequencing or assembly of small, non-real fragments. Subsampled reads were assembled using Unicycler into a final, circularised, phage genome where possible that was subsequently screened using CheckV (v1.0.1; Nayfach *et al.*, 2021). Phage genes were annotated and plotted using Pharokka (v1.3.0; Bouras *et al.*, 2023) followed by Phold (https://github.com/gbouras13/phold) together annotating functions using both sequence-  and structure-based predictions, and visualised using the pharokkaplotter function.

Genome sequences of pure phages identified were uploaded, alongside closest matches from the INPHARED database identified using MASH as part of the Pharokka pipeline (Laslett and Canback, 2004; Chen *et al.*, 2005; Bland *et al.*, 2007; Ondov *et al.*, 2016; Steinegger and Söding, 2017; McNair *et al.*, 2019; Alcock *et al.*, 2020; Cook *et al.*, 2021; P. P. Chan *et al.*, 2021; Terzian *et al.*, 2021; Shimoyama, 2022; Bouras *et al.*, 2023) were uploaded into the VIRIDIC webserver (Moraru, Varsani and Kropinski, 2020; https://rhea.icbm.uni-oldenburg.de/viridic/). In line with the recommendations from the ICTV, novel species were assigned when less than 95% intergenomic similarity to the most similar viruses was observed (Adriaenssens and Brister, 2017).

Phage genomes were compared with global *Pseudomonas* phage diversity using vConTACT3 (v3.0.0.b74; bolduc 2025) where CPL phage genomes were run alongside the vConTACT3 database (v220), the output exported as a cytoscape json file, and visualised in cytoscape v3.10.4. Nodes were filtered to include only *Pseudomonas* phages, and both singleton and small clusters of phages were arranged for clarity.

Subsequent visualisation as synteny plots was performed using Easyfig v2.2.5 (Sullivan, Petty and Beatson, 2011) with annotated phage genomes as GenBank files and pairwise BLASTn v2.15.0+ (Zhang *et al.*, 2000) between FASTA sequences generated using `blastn -query phage1.fna -subject phage2.fna -out phage1_phage2.blastn.out -outfmt 6`.  For the comparison of phage Tor and its variant capable of inhibiting DFI10.1, reads were generated in an identical manner, with SNPs being identified by running Snippy (https://github.com/tseemann/snippy) on these reads with default settings.

#### Determination of phage receptors through knockout mutant generation

Unmarked gene deletions were generated through two-step allelic exchange with a suicide vector (Hmelo *et al.*, 2015; Tong et al 2026). 500 bp sequences flanking the target genes were amplified, joined together by overlap extension PCR, and ligated into the plasmid, pEX19gm. Modified plasmids were cloned in DH5α *E. coli* and then transformed into Rho3 *E. coli* for conjugation into *P. aeruginosa* PAO1 on SOB agar (1.5% agar, 20 mM bacto-tryptone, 5 mM bacto-yeast extract, 0.5 mM NaCl, 2.5 mM KCl, 10 mM MgCl~2~)*.* Double recombinants were obtained by counter-selective plating on 100 mg/mL gentamycin and 5% *w/v* sucrose, as pEX19gm encodes gentamycin resistance (*aaC1*) and sucrose sensitivity (*sacB*). Successful mutants were distinguished from wildtype revertants by check PCRs amplifying the region surrounding the target gene. Sanger sequencing of the PCR product confirmed successful mutations. Deletions were performed on LB agar, except the deletion of *wbpL,* which was performed on MOPS-pyruvate minimal media (for 1 litre: 100 mL 10 × MOPS (400 mM MOPS buffer, 40 mM tricine, 0.1 mM FeSO~4~, 95 mM NH~4~Cl, 2.76 mM K~2~SO~4~, 5 μM CaCl~2~ and 5.25 mM MgCl~2~, adjusted to pH 7.6 with ∼330 mM KOH), 10 mL 0.132 M K~2~HPO~4~, 10 mL 20% w/v sodium pyruvate, 15 g Bacto agar), as O-antigen deficient *wbpL* mutants have previously been found to be sensitive to self-produced pyocins on nutrient-rich media (Lee *et al.*, 2015). 

Phage receptors were mostly predicted by evaluating phage host-range across the panel of receptor mutants with spot assays. 2 µL of filtrate from each phage-containing well was spotted onto a bacterial lawn of each of the panel of knockouts and incubated overnight at 37 ^o^C. Zones of lysis were characterised as either clear plaques, turbid plaques or no plaques. If a phage forms a plaque on the parent strain (PAO1∆*hsdR*) but not on a given mutant, the phage was assumed to require the deleted structure as its receptor. Phages Tor and CPL00950 had their receptors verified by spotting 2 µL of three independent serial dilutions of phage stocks onto lawns of each knockout mutant and calculating phage titre.

#### Comparison of breadth of phages CPL00272 and CPL00950

Phages were titred on *P. aeruginosa* PAO1 and adjusted to final concentrations of 10^8^, 10^7^, 10^6^, 10^4^, and 10^2^ PFU/mL in 100 μL sLB. Bacterial strains were grown to OD600 of 0.6, and diluted to create a starting OD600 reading of approximately 0.005. One bacterial strain was tested per 96-well-plate (Greiner Bio-One, Kresmünster, Austria) and each condition was tested in triplicate. Plate seals (Excel Scientific, California, USA) and incubated at 30 ^o^C for 48 hours. At hours 0, 19, 24, 43, and 48, the OD600 of each well was assessed using a Clariostar Plus Plate Reader with Stacker (BMG LabTech, Ortenberg, Germany) after 10 seconds high intensity shaking, but the read at 48 hours was discarded due to the formation of condensation on the top of the sLB control plate influencing analysis. Phage infection was assumed if a reduction in optical density of over 50% compared to untreated bacteria-only wells at 24 hours was present. 24 hours was selected to balance having enough control bacterial growth for meaningful analysis for slower strains without having faster growing strains overgrow.

#### Generation of phage stocks and PBIN

37 phage stocks, alongside several *Escherichia coli* phage T7 replicates as negative controls, were amplified by adding 5 μL phage stock into 1 mL sLB in a 2.2 mL 96-deep-well-plate (NEST Biotechnology, Jiangsu, China). To each well, 5 μL of an overnight stock of the bacterial strain each phage was isolated using was added before sealing the plate and incubating overnight at 37 ^o^C, 200 rpm. All 1 mL was distributed over six 96-well 0.45 μm MultiScreen~HTS~ HV sterile filter plates (Millipore, Merck, Darmstadt, Germany), placed over 96-well microtitre plates (Greiner Bio-One, Kremsmünster, Austria) and centrifuged at 900 × g for 15 minutes to remove bacterial cells and produce six sterile phage stock plates. Filter plates were removed, and plate lids were both taped onto the stock plates and sealed shut using Bemis^TM^ Parafilm ^TM^ M (ThermoFisher Scientific, Loughborough, UK) to limit both well-to-well contamination and evaporation of well contents. To validate phage replication, 2 μL of each well of one stock plate was spotted onto lawns of each of the isolation hosts (PAO1, RDE1, PIC15.1, PIC18.1, and PIB63.5), alongside *Escherichia coli* BW25113 to ensure no well-to-well contamination of phage into or out of T7-containing wells occurred during the phage amplification process. 

The data was first visualised by ordering phage and bacteria by breadth and susceptibility, respectively, and calculating nestedness using the NODF value generated by the `networklevel` function from the bipartitie package (v2.21). This was compared to the range of identically calculated NODF values from 10,000 Bernoulli random chance PBINs with the same likelihood of positive interactions to determine the significance of nestedness, as suggested in Weitz *et al* (2013). The data was then re-ordered by phage intergenomic relatedness and visualised using the ComplexHeatmaps package (v2.22.0), and bacteria were clustered by hierarchical clustering of phage susceptibility profiles.

#### Generation and analysis of phage-resistant bacterial isolates

To three separate volumes of 5 mL sLB, 100 μL overnight *P. aeruginosa* PAO1 and 100 μL phage stock (10^9^ PFU/mL) was added and incubated overnight (37 ^o^C, 200 rpm). The following day, 750 μL of each mixture, termed R1, R2, and R3 was added to 750 μL sterile 50% glycerol to create cryostocks for storage at -70 ^o^C. To prevent reversion or loss of the resistance phenotype, corresponding phage stock (200 μL 10^9^ PFU/mL) and 10 mM each of MgCl~2~ and CaCl~2~ were added to 30 mL LB agar when growing resistant strains. 

Assessment of phage cross resistance was performed using PAO1 strains resisting CPL00272, CPL00071, CPL00214, and CPL00950 by direct spot testing of phage stocks onto lawns of each resistant isolate. Further analysis on the interaction between CPL00071 and CPL00272 cross-resistance was performed by planktonic killing assay where isolates were grown at 37 ^o^C in 5 mL sLB with 100 μL stock phage to maintain selection pressure. When an OD600 of 0.6 was reached, the culture was diluted to OD600 = 0.1, and 5 μL was added to 100 μL final volume of sLB containing 1x10^5^ PFU/mL of CPL00272, CPL00071, or no phage. This created an approximate starting bacterial optical density of 0.005. Growth was recorded over a 4-hour period, in line with the initial virulence index method (Storms *et al.*, 2020). 

#### Transmission Electron Microscopy of Phage Particles

Electron microscopy grids (pioloform-coated 100 mesh, Agar Scientific, Rotherham, UK) were placed onto droplets of phage stock for 3 min, washed for 3 min four times using deionised water, and then placed on droplets of 2% w/v uranyl acetate for 3 min to achieve a negative stain. After excess stain was removed and air-drying, samples were imaged with a JEOL JEM 1400 transmission electron microscope at 120 kV and imaged with a digital camera (Gatan ES 1000W, Ametek, Philadelphia, USA). Imaging was performed by the Exeter Bioimaging Centre.

#### Phage One-Step Growth Curve

Exponentially growing *P. aeruginosa* PAO1 was generated by pipetting 100 μL overnight culture into 5 mL sLB. To 1 mL of this culture, phage stock was added to a final MOI of 0.01, immediately titred, and incubated for 3 minutes at 37 ^o^C to ensure adsorption. After 3 minutes, 10 μL of the phage-bacteria mix was added to 10 mL pre-warmed spent sLB to maintain nutrient consistency between flasks, 1 mL was removed as the free+adsorbed phage fraction, and returned to the 37 ^o^C water bath.

From t=10 until t = 80, sequential samples were taken every 10 minutes and titred. The free+adsorbed phage fraction was titred. The number of free phage was assessed by pelleting the free+adsorbed fraction and titreing the supernatent, and the number of adsorbed phage was calculated by resuspending the pellet in fresh 1 mL sLB, and immediately titreing the resultant stock. Due to concerns about phages binding plastic labware (Richter, 2021) and the accurate quantification needed for this experiment, all microcentrifuge tubes used were 1.5 mL Eppendorf DNA LoBind Tubes (ThermoFisher Scientific, Waltham, MA, USA) to limit phage binding.

### Table S1

| Phage     | Original Host PsA Strain | Sample Isolated From | Given Name   | Breadth on Initial Panel (%) | Predicted Receptor | Genome Length (kBp) | Most similar phage from INPHARED    | Accession Number                  |
|--------|--------|--------|--------|--------|--------|--------|--------|--------|
| CPL0056++ | PIB63.5                  | Chicken faeces       | ChickenBoo   | 4                            | O-Antigen          | 65.8                | Pseudomonas phage Qaehs             | PX060855                          |
| CPL00070  | PIB63.5                  | Countess Weir Sewage | ChickenBoo   | 60                           | Outer Core LPS     | 65.8                | Pseudomonas phage Qaehs             | To be assigned under PRJNA1257819 |
| CPL00071  | PIC15.1                  | Citizen Freshwater   | Schmoomin    | 56                           | O-Antigen          | 66.7                | Pseudomonas phage shane             | To be assigned under PRJNA1257820 |
| CPL00072  | PIC15.1                  | Countess Weir Sewage | StHubbins    | 36                           | Outer Core LPS     | 65.9                | Pseudomonas phage Qaehs             | To be assigned under PRJNA1257821 |
| CPL00075  | PIC18.1                  | Countess Weir Sewage | Hesse        | 32                           | Outer Core LPS     |                     | Pseudomonas phage sortsol           |                                   |
| CPL00113  | PAO1                     | Citizen Freshwater   | *unnamed*    | 60                           | Outer Core LPS     | 65.7                | Pseudomonas phage vB_PaeM_E217      | PV699414                          |
| CPL00161  | PAO1                     | Citizen Freshwater   | *unnamed*    | 8                            | Type IV Pilus      | 61.9                | Pseudomonas phage Churro            | PX060856                          |
| CPL00162  | PAO1                     | Citizen Freshwater   | StogurseyBug | 60                           | Untested           | 66.0                | Pseudomonas phage shane             | PV699413                          |
| CPL00173  | PAO1                     | Citizen Freshwater   | BobbyCar     | 56                           | Outer Core LPS     | 66.9                | Pseudomonas phage sortsol           | PV699415                          |
| CPL00176  | PAO1                     | Citizen Freshwater   | CHEGG        | 60                           | Outer Core LPS     | 65.9                | Pseudomonas phage DL68              | PV699416                          |
| CPL00211  | PAO1                     | Citizen Freshwater   | Pedregal     | 4                            | Type IV Pilus      | 59.1                | Pseudomonas virus Yua               | PX060858                          |
| CPL00212  | PAO1                     | Citizen Freshwater   | Citron       | 4                            | Type IV Pilus      | 58.2                | Pseudomonas phage vB_PaeS_PAO1_Ab20 | PV699417                          |
| CPL00214  | PAO1                     | Citizen Freshwater   | Stokescottia | 8                            | Type IV Pilus      | 58.2                | Pseudomonas phage ZC01              | PX060859                          |
| CPL00215  | PAO1                     | Citizen Freshwater   | Delgoamugray | 4                            | Type IV Pilus      | 58.2                | Pseudomonas phage ZC01              | PX060860                          |
| CPL00216  | PAO1                     | Citizen Freshwater   | Tinsel       | 8                            | Type IV Pilus      | 58.2                | Pseudomonas phage ZC01              | PX060861                          |
| CPL00217  | PAO1                     | Citizen Freshwater   | Hedgehog     | 12                           | Type IV Pilus      | 58.2                | Pseudomonas phage ZC01              | PX060862                          |
| CPL00252  | PAO1                     | Citizen Freshwater   | Emmi22       | 12                           | Type IV Pilus      | 61.5                | Pseudomonas phage vB_PaeS_S218      | PV699419                          |
| CPL00253  | PAO1                     | Citizen Freshwater   | *Unnamed*    | 12                           | Type IV Pilus      | 61.5                | Pseudomonas phage vB_PaeS_S218      | PX060863                          |
| CPL00254  | PAO1                     | Citizen Freshwater   | PCSewage     | 4                            | Type IV Pilus      | 61.5                | Pseudomonas phage vB_PaeS_S218      | PV699420                          |
| CPL00272  | RDE1                     | EASW                 | Tor          | 92                           | Outer Core LPS     | 65.8                | Pseudomonas phage Pa-I              | PV699421                          |
| CPL00273  | RDE1                     | EASW                 | HairPpipe    | 68                           | Outer Core LPS     | 66.5                | Pseudomonas phage sortsol           | PV699422                          |
| CPL00274  | RDE1                     | EASW                 | *unnamed*    | 68                           | Outer Core LPS     | 66.5                | Pseudomonas phage sortsol           | PV699423                          |
| CPL00275  | RDE1                     | EASW                 | GooeyTrout   | 44                           | Outer Core LPS     | 65.7                | Pseudomonas phage sortsol           | PV699424                          |
| CPL00276  | RDE1                     | EASW                 | StodgyFox    | 68                           | Outer Core LPS     | 65.7                | Pseudomonas phage Qaehs             | PV699425                          |
| CPL00277  | RDE1                     | EASW                 | *Unnamed*    | 68                           | Outer Core LPS     | 66.5                | Pseudomonas phage sortsol           | PV699426                          |
| CPL00278  | RDE1                     | EASW                 | *Unnamed*    | 88                           | Outer Core LPS     | 66.2                | Pseudomonas phage BrSP1             | PX060864                          |
| CPL00279  | RDE1                     | EASW                 | *Unnamed*    | 60                           | Outer Core LPS     | 66.2                | Pseudomonas phage sortsol           | PX060865                          |
| CPL00280  | RDE1                     | EASW                 | SlinkyKevin  | 56                           | Outer Core LPS     | 65.8                | Pseudomonas phage DL68              | PX060866                          |
| CPL00281  | RDE1                     | EASW                 | *unnamed*    | 64                           | Outer Core LPS     | 65.6                | Pseudomonas phage Qaehs             | PV699427                          |
| CPL00282  | RDE1                     | EASW                 |              | 32                           | O-antigen          |                     |                                     | Not Sequenced                     |
| CPL00283  | RDE1                     | EASW                 | HairyPottery | 84                           | Outer Core LPS     | 65.1                | Pseudomonas phage sortsol           | PX060867                          |
| CPL00285  | RDE1                     | EASW                 | FuzzyAdvisor | 64                           | Outer Core LPS     | 65.6                | Pseudomonas phage vB_PaeM_SMS29     | PX060868                          |
| CPL00286  | RDE1                     | EASW                 | *unnamed*    | 52                           | Outer Core LPS     | 65.9                | Pseudomonas phage DL68              |                                   |
| CPL00287  | RDE1                     | EASW                 | *Unnamed*    | 72                           | Outer Core LPS     | 66.4                | Pseudomonas phage sortsol           | PV699428                          |
| CPL00947  | PAO1                     | EASW                 | *Unnamed*    | 52                           | Outer Core LPS     | 66.2                | Pseudomonas phage chumba            |                                   |
| CPL00950  | PAO1                     | EASW                 | *Unnamed*    | 68                           | Outer Core LPS     | 66.5                | Pseudomonas phage sortsol           |                                   |
| CPL00973  | PAO1                     | EASW                 | *Unnamed*    | 64                           | Outer Core LPS     | 66.3                | Pseudomonas phage chumba            |                                   |

: Table S1 - Summary of phages used in this study. The original host refers to the strain of P. aeruginosa that the phage was originally isolated on. A phage's given name was provided by the citizen scientist, and its taxonomic name is a conversion of that into a binomial format that incorporates the genus predicted by the Pharokka pipeline. Purity refers to whether another phage genome could be assembled from sequencing data. The most similar phage is as predicted by the Pharokka pipeline, using MASH to find similar phages within the INPHARED database. The receptor was predicted in Tong *et al* 2026 by inability to inhibit a knockout mutant of that cellular target. EASW = Environment Agency Sewage Water; LPS = Lipopolysaccharide.

### Table S2

| Strain             | Origin | Source                                             | Phylogenomic Group | Accession    |
|-------------|-------------|--------------------|-------------|-------------|
| PAO1               | CPL    | Burn Wound/Lab Strain                              | 1                  |              |
| PIB52.5            | CPL    | Patient with Bronchiectasis                        | Unsequenced        | N/a          |
| PIB63.5            | CPL    | Patient with Bronchiectasis                        | 2                  |              |
| RDE3               | CPL    | Patient with Cystic Fibrosis                       | 2                  |              |
| RDE4               | CPL    | Patient with Cystic Fibrosis                       | 2                  |              |
| PIB58.1            | CPL    | Patient with Bronchiectasis                        | Unsequenced        | N/a          |
| PIC18.1            | CPL    | Patient with Cystic Fibrosis                       | 1                  |              |
| NCTC12903          | CPL    | Blood Culture                                      | 1                  |              |
| PIB29.5            | CPL    | Patient with Bronchiectasis                        | Unsequenced        | N/a          |
| PIB19.8            | CPL    | Patient with Bronchiectasis                        | 1                  | CP092714.1   |
| DFI4.1             | CPL    | Patient with Diabetic Foot Ulcer                   | 1                  |              |
| DFI13.1            | CPL    | Patient with Diabetic Foot Ulcer                   | 1                  |              |
| RDE1               | CPL    | Patient with Cystic Fibrosis                       | 1                  |              |
| RDE2               | CPL    | Patient with Cystic Fibrosis                       | 1                  |              |
| PIC15.1            | CPL    | Patient with Cystic Fibrosis                       | 1                  |              |
| PIC7.1             | CPL    | Patient with Cystic Fibrosis                       | 2                  |              |
| PIC21.1            | CPL    | Patient with Cystic Fibrosis                       | 1                  |              |
| PIB9.1             | CPL    | Patient with Bronchiectasis                        | 1                  |              |
| PIB1.4             | CPL    | Patient with Bronchiectasis                        | 2                  |              |
| PIB10.9            | CPL    | Patient with Bronchiectasis                        | 1                  | CP092710.1   |
| PIC1.4             | CPL    | Patient with Cystic Fibrosis                       | 1                  |              |
| PAO258             | CPL    | Patient with Cystic Fibrosis                       | 3, 4, or 5\*       |              |
| PIB50.3            | CPL    | Patient with Bronchiectasis                        | 1                  | CP092716.1   |
| PIB45.5            | CPL    | Patient with Bronchiectasis                        | 1                  |              |
| DFI10.1            | CPL    | Patient with Diabetic Foot Ulcer                   | 3, 4, or 5\*       |              |
| LESB58             | IPAP   | Patient with Cystic Fibrosis                       | 1                  | NC_011770.1  |
| LES400             | IPAP   | Patient with Cystic Fibrosis                       | 1                  | NZ_CP006982  |
| LES431             | IPAP   | Patient with Cystic Fibrosis                       | 1                  | NC_023066    |
| C3719              | IPAP   | Patient with Cystic Fibrosis                       | 1                  | MCMM00000000 |
| DK2                | IPAP   | Patient with Cystic Fibrosis                       | 1                  | NC_018080    |
| AES-1R             | IPAP   | Patient with Cystic Fibrosis                       | 1                  | MCML00000000 |
| AUS23              | IPAP   | Patient with Cystic Fibrosis                       | 1                  | MCMN00000000 |
| AUS52              | IPAP   | Patient with Cystic Fibrosis                       | 1                  | MCMK00000000 |
| AA2                | IPAP   | Patient with Cystic Fibrosis                       | 1                  | MCMJ00000000 |
| AMT 0023-30        | IPAP   | Patient with Cystic Fibrosis                       | 1                  | MCMI00000000 |
| AMT 0023-34        | IPAP   | Patient with Cystic Fibrosis                       | 1                  | MCMH00000000 |
| AMT 0060-1         | IPAP   | Patient with Cystic Fibrosis                       | 1                  | MCNB00000000 |
| AMT 0060-2         | IPAP   | Patient with Cystic Fibrosis                       | 1                  | MCNA00000000 |
| AMT 0060-3         | IPAP   | Patient with Cystic Fibrosis                       | 1                  | MCMZ00000000 |
| PAO1_panel         | IPAP   | Burn Wound Infection/Lab Strain                    | 1                  | NC_002516    |
| PA14               | IPAP   | Burn Wound Infection/Lab Strain                    | 2                  | NC_008463    |
| PAK                | IPAP   | Non Cystic Fibrosis Human Infection                | 1                  | MCMY00000000 |
| CHA                | IPAP   | Patient with Cystic Fibrosis                       | 1                  | MCMG00000000 |
| IST27 - Mucoid     | IPAP   | Patient with Cystic Fibrosis                       | 2                  | MCMX00000000 |
| IST27 - Non-Mucoid | IPAP   | Patient with Cystic Fibrosis (mutant)              | 2                  | MCMW00000000 |
| 698333S            | IPAP   | Patient with Bronchiectasis                        | 1                  | MCMV00000000 |
| 679                | IPAP   | Patient Urine                                      | 1                  | MCMU00000000 |
| 39016              | IPAP   | Patient Eye Infection                              | 2                  | MCLX00000000 |
| 2192               | IPAP   | Patient with Cystic Fibrosis                       | 1                  | NZ_CH482384  |
| NH57388A           | IPAP   | Patient with Cystic Fibrosis                       | 1                  | MCMT00000000 |
| 1709-12            | IPAP   | Non Cystic Fibrosis Human Infection                | 1                  | LZQH00000000 |
| Mil 162            | IPAP   | Burn Wound Infection                               | 2                  | MCMF00000000 |
| Jpn 1563           | IPAP   | Environmental - Lake Water                         | 1                  | MCMS00000000 |
| LMG 14084          | IPAP   | Environmental - Water                              | 2                  | MCMR00000000 |
| Pr335              | IPAP   | Environmental - Hospital                           | 1                  | MCMQ00000000 |
| U018a              | IPAP   | Patient with Cystic Fibrosis                       | 1                  | MCMP00000000 |
| CPHL 9433          | IPAP   | Environmental - Tobacco Plant                      | 2                  | MCMO00000000 |
| RP1                | IPAP   | Patient with Cystic Fibrosis                       | 1                  | LNBU00000000 |
| 15108-1            | IPAP   | Patient Infection from Intensive Care              | 2                  | MCME00000000 |
| 57P31PA            | IPAP   | Patient with Chronic Obstructive Pulmonary Disease | 1                  | MCLY00000000 |
| 13121-1            | IPAP   | Patient Infection from Intensive Care              | 1                  | MCMD00000000 |
| 39117              | IPAP   | Patient Eye Infection                              | 1                  | MCMC00000000 |
| KK1                | IPAP   | Patient with Cystic Fibrosis                       | 1                  | MCMB00000000 |
| A5803              | IPAP   | Patient with Pneumonia                             | 2                  | MCMA00000000 |
| TCBF 10839         | IPAP   | Patient with Cystic Fibrosis                       | 1                  | MCLZ00000000 |

Table S2 - Summary of bacterial isolates used in analysis. Origin refers to the collection the isolate is housed in (CPL = Citizen Phage Library; IPAP = International *P. aeruginosa* strain panel. Phylogenomic group refers to the group the strain resides in as defined in Freschi 2018 from core genome phylogeny. IPAP strain genomes acquired using accessions provided in Cullen *et al*, 2015. Strains PIB52.5, PIB58.1, and PIB29.5 are in the CPL collection but were not successfully assembled, so were not included in the phylogenomic analysis. Strains NN2, AA43, AA44 are included in the IPAP collection, but genome sequences were not available.

\*strains DFI10.1 and PAO258 fell outside groups 1 and 2 and could not be defined, but are presumed to reside in wider groups 3, 4, or 5 as defined in Freschi 2018.

### Figure S1

```{r fig.asp=0.5, fig.width=12, warning=FALSE, message=FALSE, dpi=300, out.height='5in', out.width='10in', cache=FALSE}
#| fig-cap: Unrooted (radial) phylogram produced using the equal-angle method using the data shown in figure 1C. The tree was produced using the ggtree R package, with green backgrounds representing phylogenomic group 1 and blue representing phylogenomic group 2.
#| fig-dpi: 300

nwk <- read.newick("rawdata/figure1C/core-genes-full.nwk", node.label='support')
df_tip_data <- read.csv("rawdata/figure1C/panel_strains.csv") %>% 
  mutate(Origin = ifelse(panel == "Y", "Panel Strain", "This Study")) %>% 
  select(-panel)

scale_plot <- ggtree(nwk, layout = "equal_angle") %<+% df_tip_data +
  geom_treescale(offset = 0.0005, width = 0.01, linesize = 1) +
  geom_tippoint(aes(color=Origin), shape=16, size=3) +
  scale_colour_manual(values = c("#fdb462", "#a35eb5")) +
  geom_hilight(node=69, fill="steelblue", alpha=.4) +
  geom_hilight(node=c(109,85,83,82), fill="darkgreen", to.bottom = TRUE, alpha=.4) +
  theme(plot.margin = grid::unit(c(-100, 15, -100, 15), "mm"),
        legend.text = element_text(size = 20),
        legend.title = element_text(size = 20))

scale_plot

ggsave("PDF-of-figures/figureS1.pdf", plot = scale_plot, dpi = 300)

```

### Figure S2

```{r fig.asp=0.2, fig.width=10, warning=FALSE, message=FALSE, dpi=300, out.height='2in', out.width='10in'}
#| fig-cap: Synteny plot produced using Easyfig of Tor with highly similar phage CPL00950 from the CPL.
#| fig-dpi: 300

#synteny plot

#use magick to read in the bmp produced by easyfig
synteny <- magick::image_read("rawdata/figureS2/Easyfig.bmp")
#resise the file as it is currently massive
synteny <- magick::image_scale(synteny, "900")
#add a border for labelling room
synteny <- magick::image_border(synteny, "white","105x0")
#add in labels referring to which phage is which
synteny <- magick::image_annotate(synteny, "CPL00272\n    (Tor)", size = 22, color = "#fdb462", location = "+0+0")
synteny <- magick::image_annotate(synteny, "CPL00950", size = 22, color = "#80b1d3", location = "+0+120")
#create an empty ggplot and then draw the synteny image over the top of it
figure <- ggplot() + theme_void() + draw_image(synteny)


figure
ggsave("PDF-of-figures/figureS2.pdf", plot = figure, dpi = 300)
```

### Figure S3

```{r}
#| fig-cap: "Images of phages Tor (top left, outer core LPS), CPL00950 (top middle, outer core LPS), a cocktail of Tor and CPL00950 (top right, 1:1 ratio), CPL00214 (bottom left, type IV pilus), CPL00071 (bottom middle, O-antigen), and sLB containing a non-infecting control phage was spotted onto bacterial lawns of susceptible PAO1 in panel A, or three generated resistant strains used throughout this work: B) R1, C) R2, and D) R3. All phages were at 10^8 PFU/mL when spotted."
#| fig-dpi: 300

# Image of plaques
# use magick to read in the tif produced by bioimaging
figs3 <- magick::image_read("rawdata/figureS3/figureS3.png")
#too big, have to resise
figs3 <- magick::image_scale(figs3, "1000")
#now make the figure
figs3 <- ggplot() + theme_void() + draw_image(figs3)

figs3

ggsave("PDF-of-figures/figureS3.pdf", plot = figs3, dpi = 300)

```

### Figure S4

```{r fig.asp=1.41, fig.width=30, warning=FALSE, message=FALSE, dpi=300, out.height='14.1in', out.width='10in'}
#| fig-cap: "Breadth of phages CPL00272 (Tor) and CPL00950 assessed by planktonic killing assay on the CPL bacterial panel. Inhibition was assumed if there was a 50% reduction in optical density of phage treated wells compared to bacteria-only wells for all three replicates at each titre-phage combination"
#| fig-dpi: 300

dictionary <- read.csv("rawdata/figure2B/Dictionary.csv")
strains <- read.csv("rawdata/figure2B/strains.csv")

#okay, so now creating the reformat function to do these things
reformat <- function(df){
  df <- df %>%
    select(-Content) %>%
    slice(-1) %>%
    left_join(dictionary)
  colnames(df)[2] <- "OD600"
  return(df)
}
#now need to make the function that reads in the files in the current working directory
#and adds the plate number, and the time.
produce_df <- function(file_name, time){
  df <- read.csv(file_name, header = TRUE, row.names = NULL)
  df <- reformat(df)
  file_base_name <- sub(".*(Plate \\d+).*", "\\1", basename(file_name))
  df$plate <- file_base_name
  df$time_hours <- time
  df <- df %>% left_join(strains)
  return(df)
}
#Now to go through this for timepoint 0 to make sure I can do it!
csv_files <- list.files(
  path = "rawdata/figure2B/0 hour/", 
  pattern = "\\.CSV$", 
  full.names = TRUE
)
#use purrr::map_df to apply the reformat_csv function to each CSV file and bind them together
df_0 <- map2(csv_files, c(0), produce_df)

#repeat for the other timepoints and then cbind them together
#t19
csv_files <- list.files(
  path = "rawdata/figure2B/19 hour/", 
  pattern = "\\.CSV$", 
  full.names = TRUE
)
#use purrr::map_df to apply the reformat_csv function to each CSV file and bind them together
df_19 <- map2(csv_files, c(19), produce_df)


#t24
csv_files <- list.files(
  path = "rawdata/figure2B/24 hour/", 
  pattern = "\\.CSV$", 
  full.names = TRUE
)
#use purrr::map_df to apply the reformat_csv function to each CSV file and bind them together
df_24 <- map2(csv_files, c(24), produce_df)


#t43
csv_files <- list.files(
  path = "rawdata/figure2B/43 hour/", 
  pattern = "\\.CSV$", 
  full.names = TRUE
)
#use purrr::map_df to apply the reformat_csv function to each CSV file and bind them together
df_43 <- map2(csv_files, c(43), produce_df)

#t48
csv_files <- list.files(
  path = "rawdata/figure2B/48 hour/", 
  pattern = "\\.CSV$", 
  full.names = TRUE
)
#use purrr::map_df to apply the reformat_csv function to each CSV file and bind them together
df_48 <- map2(csv_files, c(48), produce_df)


#produce a df_all for all the data
df_produce <- function(df_list){
  final_df <- Reduce(full_join, df_list)
  return(final_df)
}
#bind them all together
df_0 <- df_produce(df_0)
df_19 <- df_produce(df_19)
df_24 <- df_produce(df_24)
df_43 <- df_produce(df_43)
df_48 <- df_produce(df_48)
df_all <- rbind(df_0,df_19,df_24,df_43,df_48)
#remove the annoying X column
df_all <- df_all %>% select(-X)
#we are not interested in a cocktail between the two, although I have that data, its not relevant anymore
df_all <- df_all %>% filter(Phage != "Cocktail") %>% mutate(Phage = ifelse(Phage == "CPL00272", "Tor", Phage))
#now as I messed up and added phage to B07 to B11, remove these! It also 
#turns out the OD600 row is a character, changing that to numeric!
dodgy <- c("B06", "B07", "B08", "B09", "B10", "B11")
df_all <- df_all %>%
  filter(!(Well %in% dodgy)) %>%
  mutate(OD600 = as.numeric(OD600))

#check the blanks for each plate!
# ggplot(filter(df_all, bacteria == "LB"), aes(x = time_hours, y = OD600, colour = plate)) +
#   geom_point()
#it seems that hour 48 is weird in terms of growth, will remove this data! This is likely due to condensation build up on the plate for some wells. Other timepoint blanks look okay
df_all <- df_all %>%
  filter(time_hours != 48)


#now we want to identify the blank dataframe
df_blanks <- df_all %>%
  filter(bacteria == "LB") %>%
  group_by(time_hours) %>%
  reframe(mean_blk_od = mean(OD600))
#now we will normalise all the ODs
df_all <- df_all %>%
  left_join(df_blanks) %>%
  mutate(norm_od = OD600 - mean_blk_od)
#calculate summary statistics for each of these
df_all_means <- df_all %>%
  group_by(bacteria, Phage, Concentration, time_hours) %>%
  mutate(mean_od_timepoint = mean(norm_od), stdev = sd(norm_od))
#fixing some buggy things regarding the titre being in form 6e+001 or something
df_all <- df_all %>%
  mutate(titre = log10(Concentration)) %>%
  mutate(titre = ifelse(is.infinite(titre) == TRUE, 0, titre)) %>%
  mutate(titre = factor(titre))
#isolate the controls (no phage) and calculate summary statistics
df_controls_only <- df_all %>%
  filter(Phage == "Control") %>%
  group_by(time_hours, bacteria) %>%
  reframe(mean_od = mean(norm_od), stdev = sd(norm_od)) %>%
  mutate(titre = 0)
#remove the control form the original dataframe, as these are clarified elsewhere
df_all <- df_all %>%
  filter(Phage != "Control")

df_means <- df_all %>%
  group_by(Phage, titre, time_hours, bacteria) %>%
  reframe(mean_od = mean(norm_od), stdev = sd(norm_od))

#to get raw numbers of reduction percentage of the phage at each timepoint we can left join in a means column from df_means
#then calculate reduction pecentage by each phage at each timepoint of interest:

df_reduction <- df_all %>%
  left_join(select(df_controls_only, time_hours, bacteria, mean_od)) %>%
  rename(mean_control_od = mean_od) %>%
  mutate(red_perc = 100*(mean_control_od-norm_od)/mean_control_od)

df_reduction_plot <- df_reduction %>%
  group_by(time_hours, Phage, bacteria, titre) %>%
  reframe(mean_red_perc = mean(red_perc), stdev = sd(red_perc))

df_reduction_plot_modified <- df_reduction_plot %>%
  filter(bacteria != "LB") #remove LB as this one is too noisy for this to make sense

means <- ggplot(filter(df_means, titre == 2 | titre == 4 | titre == 8), aes(x = time_hours, y = mean_od, colour = titre)) +
  geom_line() +
  geom_errorbar(aes(ymin = mean_od - stdev, ymax = mean_od + stdev), width = 0.1) +
  geom_line(data = df_controls_only, aes(x = time_hours, y = mean_od), colour = "black") +
  geom_errorbar(data = df_controls_only, aes(ymin = mean_od - stdev, ymax = mean_od + stdev), width = 0.1, colour = "black") +
  geom_point() +
  geom_point(data = df_controls_only, aes(x = time_hours, y = mean_od), colour = "black") +
  scale_colour_manual("Log10\nPhage Titre\n(PFU/mL)", values = okabe) +
  facet_wrap(Phage ~ bacteria, nrow = 4) +
  scale_x_continuous("Time (hours)", breaks = seq(0,40,20)) +
  scale_y_continuous("Mean OD600 (n = 3)", limits = c(-0.3, 2)) +
  theme_cowplot(28)

#at some timepoints, the bacterial growth is too low to actually create a meaningful comparison. 
#These bacteria are PIB63.5, PIB50.3, PIC1.4, PIB52.4, RDE3, RDE4 cause particular issues with this
#every bacteria barring 52.5 has substantial growth at 24 hours, so might make the most meaningful comparison to use this timepoint
  
overall_red <- ggplot(filter(df_reduction_plot_modified, time_hours == 24), aes(x = titre, y = mean_red_perc, fill = titre)) +
  geom_col(position = position_dodge()) +
  scale_x_discrete("Log10 Phage Titre (PFU/mL)") +
  scale_y_continuous("Mean Reduction Percentage (n = 3, +- SD)", limits = c(-300, 250), breaks = seq(-300, 200, 100)) +
  geom_errorbar(aes(ymin = mean_red_perc - stdev, ymax = mean_red_perc + stdev)) +
  facet_wrap(Phage ~ bacteria, nrow = 5) +
  scale_fill_manual("Log10\nPhage Titre\n(PFU/mL)", values = okabe) +
  geom_hline(yintercept = 50, colour = "grey", linetype = "dotdash") +
  theme_cowplot(28)

figure <- plot_grid(means, overall_red, nrow = 2, labels = "AUTO")

figure
ggsave("PDF-of-figures/figureS4.pdf", plot = figure, dpi = 300)
```

### Figure S5

```{r fig.asp=1, fig.width=12, warning=FALSE, message=FALSE, dpi=300, out.height='8.27in', out.width='8.27in'}
#| fig-cap: "Growth over four hours of a phage-susceptible wild type (WT) and three independantly isolated phage-resistant isolates (R1, R2, and R3) to Tor, CPL00071, and no phage."
#| fig-dpi: 300

#cross resistance

#First, read in the data
df_4h <- read.csv("rawdata/figureS5/First_four_hours.csv")
dictionary = read.csv("rawdata/figureS5/dictionary.csv")

#This function reformats each individual CSV to produce a long format that is usable. Remove rows with NAs as some tecan runs mess up.
reformat <- function(df) {
  colnames(df)[1] <- "Well"
  new_names <- sapply(colnames(df)[-1], function(x) substr(x, 2, nchar(x) - 1))
  colnames(df)[-1] <- new_names
  df_long <- pivot_longer(df, cols = -Well, names_to = "Time_Hours", values_to = "values")
  df_long$Time_Hours <- as.numeric(as.character(df_long$Time_Hours))
  df_long <- mutate(df_long, Time_Hours = round(Time_Hours / 3600 / 0.25) * 0.25)
  df_long <- left_join(df_long, dictionary)
  df_long <- df_long %>% na.omit()
  return(df_long)
}

#Run the reformatting function on the dataframe and remove protector wells that are just to limit within-well evaporation
df_4h_reformatted <- df_4h %>% 
  reformat() %>%
  filter(phage != "Protector")

#Extract blank wells
df_4h_blk <- df_4h_reformatted %>%
  filter(phage == "BLK") %>%
  group_by(phage, Time_Hours) %>%
  reframe(mean_blk_od = mean(values)) %>%
  rename(blk_od = mean_blk_od) %>%
  select(-phage)

#Normalise measurements to blanks and extract whether each well containins resistant strains and to which phage.
df_4h_reformatted <- df_4h_reformatted %>%
  left_join(df_4h_blk) %>%
  mutate(norm_od = values - blk_od) %>%
  filter(phage != "BLK") %>%
  filter(bacteria != "None") %>% 
  mutate(replicate = str_extract(bacteria, ".{2}$")) %>%
  mutate(resist = paste(c(str_extract(bacteria, "^[^_]+")), "Resistant"))%>%
  mutate(resist = ifelse(resist == "PAO1 Resistant", "No Phage Resistance", resist))

#make sure we include that CPL00272 is called Tor
df_4h_plotting <- df_4h_reformatted %>%
  mutate(
    phage = ifelse(phage == "CPL00272", "Tor", phage),
    bacteria = ifelse(bacteria == "CPL00272_R1", "Tor_R1",
                      ifelse(bacteria == "CPL00272_R2", "Tor_R2",
                             ifelse(bacteria == "CPLOO272_R3", "Tor_R3", bacteria))),
    resist = ifelse(resist == "CPL00272 Resistant", "Tor Resistant", resist)
  ) %>%
  mutate(
    phage = factor(phage, levels = c("No Phage", "Tor", "CPL00071")),
    resist = factor(resist, levels = c("No Phage Resistance", "Tor Resistant", "CPL00071 Resistant"))
  )

#Plot it
figure <- ggplot(df_4h_plotting, aes(x = Time_Hours, y = norm_od, group = Well)) +
  geom_line(aes(colour = replicate)) +
  facet_grid(phage~resist) +
  scale_x_continuous("Time (Hours)") +
  scale_y_continuous("Normalised OD600") +
  scale_colour_manual("Biological\nReplicate", values = c("#fb8072", "#80b1d3", "#fdb462", "#d9d9d9")) +
  theme_cowplot(20)

figure
ggsave("PDF-of-figures/figureS5.pdf", plot = figure, dpi = 300)


#now to run a fairly complicated linear regression on this. This whole section will be commented out so that we don't run each time and do not want any outputs directly in the manuscript.

#need a quantifiable measure so will use virulence index to actually do this. Will calculate that now as in Storms et al 2020.

# #calculate AUC of each well
# df_stats <- df_4h_reformatted %>%
#   select(Well, phage, Time_Hours, norm_od, replicate, resist) %>%
#   group_by(Well, phage, replicate, resist) %>%
#   reframe(area = AUC(x=Time_Hours, y=norm_od, method='trapezoid'))
# #Identify AUC of each well for non-infected controls
# df_controls <- df_stats %>%
#   filter(phage == "No Phage") %>%
#   filter(resist == "No Phage Resistance") %>%
#   group_by(phage) %>%
#   reframe(mean_cntrl_auc = mean(area)) %>%
#   select(mean_cntrl_auc) %>%
#   as.numeric()
# #leftjoin this back in and calculate virulence index
# df_stats <- df_stats %>%
#   mutate(control = df_controls) %>%
#   mutate(Vp = 1 - (area/control))
# #set everything up in the dataframe to make linear regressions use the correct baselines etc
# df_stats <- df_stats %>%
#   mutate(resist = factor(resist, levels = c("No Phage Resistance","CPL00071 Resistant", "CPL00272 Resistant"))) %>%
#   mutate(phage = factor(phage, levels = c("No Phage", "CPL00071", "CPL00272"))) %>%
#   mutate(replicate = factor(replicate, levels = c("WT", "R1", "R2", "R3"))) %>%
#   select(-c(Well, area, control))
# 
# #cant account for multiple independant replicates due to high multicolinearity, so the model is this:
# model_Xres <- lm(Vp ~ phage*resist, data = df_stats)
# 
# # plot(model_Xres) # not amazing but gives a good idea
# # summary(model_Xres)
# #this says that compared to the wild type, the phages (CPL00071 and CPL00272) have an increased virulence. It also suggests that resistant bacteria have some level of reduced growth rate. This isn't the question though and not the one reported in the paper. The actual question of interest is does CPL00272 confer resistance to 71 i.e. is the virulence index of CPL00071 in CPL00272-resistant strains different to adding 272 to these strains?
# 
# df_1 <- df_stats %>%
#   filter(resist == "CPL00272 Resistant") %>%
#   filter(phage != "No Phage") %>%
#   mutate(phage = factor(phage, levels = c("CPL00272", "CPL00071")))
# 
# model_q1 <- lm(Vp ~ phage, data = df_1)
# summary(model_q1)
# plot(model_q1)
# 
# #fine - no significant difference between CPL00272 and CPL00071 in these
# #two hosts. Therefore, CPL00272 conveys cross resistance to 71
# 
# #2 ---------
# #does 71 confer resistance to 272, i.e. is the virulence index of 272 in 71-resistant
# #strains different to adding 71 to these strains?
# 
# df_2 <- df_stats %>%
#   filter(resist == "CPL00071 Resistant") %>%
#   filter(phage != "No Phage") %>%
#   mutate(phage = factor(phage, levels = c("CPL00071", "CPL00272")))
# 
# model_q2 <- lm(Vp ~ phage, data = df_2)
# summary(model_q2)
# plot(model_q2)

#fine - no significant difference between CPL00272 and CPL00071 in these
#two hosts. Therefore, CPL00071 conveys resistance to 272.

#model is a bit less than ideal here, particlularly regarding the QQ plot,
#but given the fact this looks exactly like the few P71-resistant
#strains that didnt grow, left in as non-significant.
```

### Figure S6

```{r fig.asp=1.41, fig.width=12, warning=FALSE, message=FALSE, dpi=300, out.height='11.69in', out.width='8.27in'}
#| fig-cap: Phage-Bacteria Interaction Network (PBIN) of 37 tested phage on 25 P. aeruginosa host strains. A) Bacteria were grown with or without phage for 24 hours at 30 degrees, and inhibition was assumed if a greater than 50% reduction in the area under the untreated bacterial control was observed. Phages are ordered by intergenomic similarity as determined by VIRIDIC above, and bacteria ordered by heirachical clustering. B) Nestedness of 10,000 randomly generated bernoulli-random chance PBINs where the probablility of phage inhibition is equal to that of the PBIN in panel A. The NODF value of the PBIN generated here is shown by the red vertical line.
#| fig-dpi: 300 
dictionary = read.csv("rawdata/figure2C/Dictionary.csv")

#when I read in each individual dataset, I will be using a function to format it into a specific way, and then join them all together.
#will be defining that function here. 

csv_files <- list.files(
  path = "rawdata/figure2C/GrowthCurveData/", 
  pattern = "\\.csv$", 
  full.names = TRUE
)

#This function reformats each individual CSV to produce a long format that is usable. Remove rows with NAs as some tecan runs may mess up.
reformat <- function(df) {
  colnames(df)[1] <- "Well"
  new_names <- sapply(colnames(df)[-1], function(x) substr(x, 2, nchar(x) - 1))
  colnames(df)[-1] <- new_names
  df_long <- pivot_longer(df, cols = -Well, names_to = "Time_Hours", values_to = "values")
  df_long$Time_Hours <- as.numeric(as.character(df_long$Time_Hours))
  df_long <- mutate(df_long, Time_Hours = round(Time_Hours / 3600 / 0.25) * 0.25)
  df_long <- left_join(df_long, dictionary)
  df_long <- df_long %>% na.omit()
  return(df_long)
}

#This function then applies each csv in the working directory folder to the previous function
#and adds a column that uses the filename to state which strain is in use
reformat_csv <- function(file_name) {
  df <- read.csv(file_name, header = TRUE, row.names = NULL)
  df <- reformat(df)
  file_base_name <- sub("_.*", "", basename(file_name))
  df$bacteria <- file_base_name
  return(df)
}

#use purrr::map_df to apply the reformat_csv function to each CSV file and bind them together
df_all <- map_df(csv_files, reformat_csv)

#it turns out well H12 is contaminated with phage and H9 will be excluded.
df_all <- df_all %>% filter(Well != "H12") %>% filter(Well != "H9")

#going to need to calculate virulence index for each phage, in each host
df_summary <- df_all %>% group_by(Time_Hours, bacteria, phage) %>% 
  reframe(bacteria = bacteria, phage = phage, Time_Hours = Time_Hours, mean_od = mean(values))

#extract blank dataframe and left_join it
blks_df = df_summary %>% filter(phage == "BLK") %>% group_by(Time_Hours) %>% summarise(mean_blk_od = mean(mean_od))
df_summary = df_summary %>% left_join(blks_df, by = c("Time_Hours")) %>% mutate(norm_od = mean_od - mean_blk_od)

#calculate the AUC of the controls, and join this into df_summary
#will fire an equal number of warnings as bacteria it collapses each value into a single column
df_virulence = df_summary %>% filter(phage == "No Phage") %>% group_by(bacteria, Time_Hours) %>% 
  reframe(bacteria = bacteria, mean_ctrl_od = mean(norm_od)) %>% group_by(bacteria) %>%
  reframe(mean_ctrl_area = (AUC_trap=AUC(x=Time_Hours, y=mean_ctrl_od, method='trapezoid')))
df_summary = df_summary %>% left_join(df_virulence)

#calculate virulence index for each well in df_summary
#will fire twice number of warnings as bacteria it collapses each value into a single column
df_summary = df_summary %>% group_by(bacteria, phage) %>% 
  reframe(meanAUC = AUC(x=Time_Hours, y=norm_od, method='trapezoid'), mean_ctrl_area = mean_ctrl_area) %>%
  unique() %>% reframe(bacteria = bacteria, phage = phage, Vp = (1 - (meanAUC/mean_ctrl_area)))

#now to create the PBIN, I need to work out if each phage infects a bacteria:
#if Vp > 0.5, then the phage infects the bacteria. 
df_summary <- df_summary %>% mutate(Infection = ifelse(Vp > 0.5, 1, 0))

#now create the PBIN!!
PBIN <- df_summary %>% select(bacteria, phage, Infection) %>%
  pivot_wider(names_from = phage, values_from = Infection)

#As I am generally interested, lets do a data comparison to calculate host range % of each phage
#Need to sum up each column of the PBIN and divide by the number of rows
PBIN_summary <- PBIN %>% select(-bacteria)
n_rows <- nrow(PBIN)
sums <- colSums(PBIN_summary) / n_rows
PBIN_summary <- data.frame(phage = colnames(PBIN_summary), Percentage_Infected = sums*100)
PBIN_summary <- PBIN_summary %>% arrange(Percentage_Infected)

#Going to make a preliminary heatmap plot of Pseudomonas phages as rows against the hosts as columns, ranked from top to bottom by the row sum of each phage (i.e the most virulent at the top)
#Got to first get PBIN into a decnt format for this, need to be pivoted longer.
#Will need X axis of bacteria, Y axis of phage, and a third column for if the interaction happens.

#need to extract the order of phage infectivity from the PBIN_summary df
Phage_order <- PBIN_summary %>% filter(phage != "BLK")
Phage_order <- unlist(as.vector(select(Phage_order[order(Phage_order$Percentage_Infected),], -Percentage_Infected)))

bacteria_order <- PBIN %>% mutate(Promiscuity = rowSums(across(where(is.numeric)))) %>% select(bacteria, Promiscuity)
bacteria_order <- c(unlist(as.vector(select(bacteria_order[order(-bacteria_order$Promiscuity),], -Promiscuity))))

exclude_impure_and_sedgewick <- c("CPL00934", "CPL00937", "CPL00951", "CPL00954", "CPL00933", "CPL00935", "CPL00953", "CPL00943", "CPL00936", "CPL00938", "CPL00944", "CPL00941", "CPL00939",  "CPL00940", "CPL00942", "CPL00946", "CPL00949", "CPL00952", "CPL00165")

Heatmap_PBIN <- pivot_longer(PBIN, cols = -bacteria, names_to = "Phage", values_to = "Infection") %>%
  mutate(Infection = factor(Infection, levels = c("1", "0"))) %>% 
  mutate(Phage = factor(Phage, levels = Phage_order)) %>%
  filter(Phage != "T7") %>% 
  mutate(bacteria = factor(bacteria, levels = bacteria_order)) %>%
  filter(!Phage %in% exclude_impure_and_sedgewick) 


Heatmap_plot <- ggplot(Heatmap_PBIN, aes(x = bacteria, y = Phage, fill = Infection)) +
  geom_tile(colour = "white", lwd = 1, linetype = 1) +
  scale_fill_manual(values = c("1" = "#bc80bd", "0" = "#8dd3c7"), labels=c('Inhibtiion', 'No Inhibition')) +
  scale_y_discrete('Phage') +
  scale_x_discrete("Pseudomonas aeruginosa Strain") +
  labs(fill = "Phage Infection") +
  theme_classic(base_size=20) +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_rect(colour = "black")) +
  theme(axis.text.x = element_text(angle = 90))


#This is for the stats for nestedness! Following the method in Weitz et al 2013

#I'll need the PBIN in a state that is like the PBIN file, except we cant be having a
#row for blanks and a row for no phage
PBIN_analysis <- PBIN %>%
  select(-c("No Phage", "BLK")) %>%
  column_to_rownames(var='bacteria')

#calculate the NODF value

# networklevel(PBIN_analysis, index = "NODF") #commented out as it will appear in rendered document if I do not
#NODF value is 79.82427

#okay, so need to create a list of numbers with the nestedness values of many
#null models. I dont want to do a for loop, but it seems to make the most sense? 

#how I determined the number of positive interactions:
# Heatmap_count = Heatmap_PBIN %>%
#   filter(Infection == 1) %>%
#   nrow()


#I need to create a function that creates the dataframe I want for each null model.
#There are a total of 25*38=950 interactions, of which 432 are positive
#This gives a positive rate of 432/950 = 0.4547368
#Enables me to generate null models using the bernoulli method


#dummy variables that match the dimensions of the PBIN
bacterial_names <- rownames(PBIN_analysis)
phage_names <- colnames(PBIN_analysis)

#create a series of lists that contain 1s and 0s with the correct probability
sequences <- function(x){
  output = list()
    for (i in 1:x){
    values <- rbinom(n = 25, size = 1, prob = 0.4547368)
    output[[i]] <- values
    }
  return(output)
}

#run the above function, create the dataframe, perform the NODF calculation
null_model <- function(x){
  distribution_of_NODF <- vector()

  for (i in 1:x){
    test <- sequences(58)
    test_df <- as.data.frame(do.call(cbind, test))
    row.names(test_df) <- bacterial_names
    colnames(test_df) <- phage_names
    test_df <- test_df %>% select(-T7)
    result <- unname(networklevel(test_df, index = "NODF"))
    distribution_of_NODF <- c(distribution_of_NODF, result)
  }
  return(distribution_of_NODF)
}
#calculate random chance nestedness values

playing_output <- null_model(10000) #commented out as takes a while
#range(playing_output) #range was 39.98116 to 51.86719. Note these values will change slightly from run to run as all randomly generated.
histogram_plot <- ggplot(data.frame(playing_output), aes(x = playing_output)) +
  geom_histogram(binwidth = 1, fill = "grey70", color = "black") +
  theme_cowplot(20) +
  geom_vline(xintercept = 79.82427, colour = "red") +
  labs(x = "NODF value", y = "Count",
       title = "Distribution of Random PBIN NODF Values")

figure <- plot_grid(Heatmap_plot, histogram_plot, nrow = 2, rel_heights = c(4, 1), labels = "AUTO")

#as my PBIN NODF value was 79.82427 and of 10,000 random PBINs there was not a single one that hit that value, then my PBIN is significantly nested with a p-value less than 0.0001.

figure
ggsave("PDF-of-figures/figureS6.pdf", plot = figure, dpi = 300)
```

### Figure S7

```{r fig.asp=0.9, fig.width=20, warning=FALSE, message=FALSE, dpi=300, out.height='7.44in', out.width='8.27in'}
#| fig-cap: "Genomic features of CPL00272 (Tor) predicted using the Pharokka pipeline, followed by Phold to provide sequence- and structure-based predictions of protein function. The output image from PharokkaPlotter was edited in Inkscape for clarity."
#| fig-dpi: 300

#Genomic features of CPL00272
#use magick to read in the png produced by phold and edited in inkskape
genome_plot <- magick::image_read("rawdata/figureS7/phold-plot-edited.png")
#too big again, need to rescale
#genome_plot <- magick::image_scale(genome_plot, "10000")
#now make the figure
figure <- ggplot() + theme_void() + draw_image(genome_plot)

figure
ggsave("PDF-of-figures/figureS7.pdf", plot = figure, dpi = 300)

```

### Figure S8

```{r fig.asp=0.9, fig.width=20, warning=FALSE, message=FALSE, dpi=300, out.height='7.44in', out.width='8.27in'}
#| fig-cap: "Intergenomic similarity as calculated using VIRIDIC between Tor, CPL00950, and the three most used Pseudomonas phages in the BT100 study (14-1: NC_011703.1; PNM: OP292288.1; PT07: OQ850183.1). Text represents the percentage intergenomic similarity."
#| fig-dpi: 300

#Comparison of CPL00272 to other prominent Viruses
belgium_phage <- read.csv("rawdata/figureS8/VIRIDIC.csv") %>%
  pivot_longer(!genome, names_to = "comparison", values_to = "similarity") %>%
  mutate(comparison = ifelse(comparison == "X14.1", "14-1", comparison)) %>%
  mutate(genome = ifelse(genome == "CPL00272", "Tor", genome)) %>%
  mutate(comparison = ifelse(comparison == "CPL00272", "Tor", comparison))

#isolate the order of the VIRIDIC output to ensure final graph is the same order (matching VIRIDIC's clustering)
phage_order <- belgium_phage %>%
  select(genome) %>%
  unique() %>%
  as.vector() %>%
  unlist()

#set the order of the phage genomes for ggplot
belgium_phage <- belgium_phage %>%
  mutate(genome = factor(genome, levels = phage_order)) %>%
  mutate(comparison = factor(comparison, levels = phage_order))

#plot it as above
phage_belgium_plot <- ggplot(belgium_phage, aes(x = genome, y = comparison, fill = similarity)) +
  geom_tile() +
  geom_text(aes(label = similarity), size = 16) +
  scale_y_discrete("Phage Genome") +
  scale_x_discrete("Phage Genome") +
  scale_fill_gradient2("Whole Genome\nIntergenomic\nSimilarity (%)", limits = c(0,100), low = "#ffffff", mid = "#d8d7f5", midpoint = 60, high = "#0400ff") +
  theme_cowplot(28) +
  theme(axis.text.x = element_text(angle = 90))
phage_belgium_plot

ggsave("PDF-of-figures/figureS8.pdf", plot = phage_belgium_plot, dpi = 300)
```

### Figure S9

```{r fig.asp=1, fig.width=20, warning=FALSE, message=FALSE, dpi=300, out.height='11.69in', out.width='11.69in'}
#| fig-cap: "Growth curves of PAO1 either uninfected (n = 2) or in the presence of 10^7 PFU/mL of either phage Tor (n = 4), phage PNM (n = 4), or a 1:1 cocktail (n=2) of Tor and PNM."
#| fig-dpi: 300

#read in the synogram data
row_content <- read.csv("rawdata/figureS9/row_data.csv")
col_content <- read.csv("rawdata/figureS9/column_data.csv")

#This function will reformat an individual .CSV into the correct format by modifying the columns, reformatting the time
#adding in the dictionary that tells us what is in each well, and removes NAs for dodgy values
reformat <- function(df_test) {
  df_test <- df_test %>% select(-Well)
  df_test[1,1] <- "Well"
  #names(df_test) <- lapply(df_test[1, ], as.character)
  df_test <- df_test[-1,] 
  df_test <- rename(df_test, "Time_Hours" = "X")
  
  df_test <- pivot_longer(df_test, cols = -Time_Hours, names_to = "Well", values_to = "values") %>%
    filter(Time_Hours != "Time")
  
  df_test <- df_test %>%
    mutate(
      hours = as.numeric(str_extract(Time_Hours, "\\d+(?=\\s*h)")),
      minutes = as.numeric(str_extract(Time_Hours, "\\d+(?=\\s*min)")),
      hours = replace_na(hours, 0),
      minutes = replace_na(minutes, 0),
      Time_seconds = (hours * 3600) + (minutes * 60)
    ) %>%
    select(-hours, -minutes)  # Remove temporary columns
  
  df_test <- df_test %>%
    mutate(
      Row = str_extract(Well, "^[A-Z]"),
      Column = as.numeric(str_extract(Well, "\\d+$"))
    )
}


#This function now will take each of the filenames in the folder, and for each it'll take in
#the file name, run the reformatting function above, add in a column with the plate name
#(taken from the file name) 

produce_df <- function(file_name){
  df <- read.csv(file_name, header = TRUE, row.names = NULL)
  df <- reformat(df)
  file_base_name <- sub("\\.\\w+$", "", basename(file_name))
  df$plate <- file_base_name
  return(df)
}

#get the list of all the csv files in the folder
csv_files <- list.files(
  path = "rawdata/figureS9/", 
  pattern = "\\.CSV$", 
  full.names = TRUE
)

#use purrr::map_df to apply the produce_df function to each CSV file and bind them together
df_all <- map(csv_files, produce_df)

#df in list form, this isnt ideal, need a concatenated one
df_list <- function(df_list){
  final_df <- Reduce(full_join, df_list)
  return(final_df)
}
df_all <- df_list(df_all)

df_all <- left_join(df_all, row_content) %>% left_join(col_content) %>% na.omit() %>%
  mutate(Row = factor(Row, levels = c("A","B","C","D","E","F","G"))) %>%
  mutate(Column = factor(Column, levels = seq(2,10,1)))

temporary_plot <- ggplot(df_all, aes(x = Time_seconds, y = values, group = Well)) +
  geom_line() +
  facet_grid(plate+Row~Column)

#data handling to make the rest of it make sense
df_all_analysis <- df_all %>%
  mutate(values = as.numeric(values)) %>%
  filter(Time_Hours != "")

#extract blanks
df_blanks <- df_all_analysis %>%
  filter(bacteria == "None") %>%
  filter(phage == "No Phage") %>%
  group_by(Time_seconds, plate) %>%
  mutate(mean_blk_od = mean(values)) %>%
  select(Time_seconds, plate, mean_blk_od)

#normalise od600 and set time to be in hours
df_all_analysis <- df_all_analysis %>%
  left_join(df_blanks) %>%
  mutate(norm_od = values - mean_blk_od) %>%
  filter(bacteria != "None") %>%
  mutate(Time_Hours = Time_seconds/(60*60)) %>%
  mutate(phage = factor(phage, levels = c("No Phage", "Tor", "PNM", "Cocktail"))) 

#plot all the growth curves
figure <- ggplot(df_all_analysis, aes(x = Time_Hours, y = abs(norm_od), group = Well, colour = phage)) +
  geom_line() +
  facet_wrap(~bacteria, nrow = 5) +
  scale_x_continuous("Time (Hours)", breaks = seq(0,48, 12)) +
  scale_y_continuous("Normalised OD600") +
  scale_colour_manual("Phage", values = okabe) +
  theme_cowplot(28)

figure
ggsave("PDF-of-figures/figureS9.pdf", plot = figure, dpi = 300)

```

### Figure S10

```{r fig.asp=1.41, fig.width=12, warning=FALSE, message=FALSE, dpi=300, out.height='11.69in', out.width='8.27in'}
#| fig-cap: "Mean phage local virulence based on uninfected (n = 2) PAO1 in the presence of 10^7 PFU/mL of either phage Tor (n = 4), phage PNM (n = 4), or a 1:1 cocktail (n=2) of Tor and PNM calculated over the 48-hour period."
#| fig-dpi: 300

#read in the synogram data
row_content <- read.csv("rawdata/figureS10/row_data.csv")
col_content <- read.csv("rawdata/figureS10/column_data.csv")

#This function will reformat an individual .CSV into the correct format by modifying the columns, reformatting the time
#adding in the dictionary that tells us what is in each well, and removes NAs for dodgy values
reformat <- function(df_test) {
  df_test <- df_test %>% select(-Well)
  df_test[1,1] <- "Well"
  #names(df_test) <- lapply(df_test[1, ], as.character)
  df_test <- df_test[-1,] 
  df_test <- rename(df_test, "Time_Hours" = "X")
  
  df_test <- pivot_longer(df_test, cols = -Time_Hours, names_to = "Well", values_to = "values") %>%
    filter(Time_Hours != "Time")
  
  df_test <- df_test %>%
    mutate(
      hours = as.numeric(str_extract(Time_Hours, "\\d+(?=\\s*h)")),
      minutes = as.numeric(str_extract(Time_Hours, "\\d+(?=\\s*min)")),
      hours = replace_na(hours, 0),
      minutes = replace_na(minutes, 0),
      Time_seconds = (hours * 3600) + (minutes * 60)
    ) %>%
    select(-hours, -minutes)  # Remove temporary columns
  
  df_test <- df_test %>%
    mutate(
      Row = str_extract(Well, "^[A-Z]"),
      Column = as.numeric(str_extract(Well, "\\d+$"))
    )
}


#This function now will take each of the filenames in the folder, and for each it'll take in
#the file name, run the reformatting function above, add in a column with the plate name
#(taken from the file name) 

produce_df <- function(file_name){
  df <- read.csv(file_name, header = TRUE, row.names = NULL)
  df <- reformat(df)
  file_base_name <- sub("\\.\\w+$", "", basename(file_name))
  df$plate <- file_base_name
  return(df)
}

#get the list of all the csv files in the folder
csv_files <- list.files(
  path = "rawdata/figureS10/", 
  pattern = "\\.CSV$", 
  full.names = TRUE
)

#use purrr::map_df to apply the produce_df function to each CSV file and bind them together
df_all <- map(csv_files, produce_df)

#df in list form, this isnt ideal, need a concatenated one
df_list <- function(df_list){
  final_df <- Reduce(full_join, df_list)
  return(final_df)
}
df_all <- df_list(df_all)

df_all <- left_join(df_all, row_content) %>% left_join(col_content) %>% na.omit() %>%
  mutate(Row = factor(Row, levels = c("A","B","C","D","E","F","G"))) %>%
  mutate(Column = factor(Column, levels = seq(2,10,1)))

temporary_plot <- ggplot(df_all, aes(x = Time_seconds, y = values, group = Well)) +
  geom_line() +
  facet_grid(plate+Row~Column)

#data handling to make the rest of it make sense
df_all_analysis <- df_all %>%
  mutate(values = as.numeric(values)) %>%
  filter(Time_Hours != "")

#extract blanks
df_blanks <- df_all_analysis %>%
  filter(bacteria == "None") %>%
  filter(phage == "No Phage") %>%
  group_by(Time_seconds, plate) %>%
  mutate(mean_blk_od = mean(values)) %>%
  select(Time_seconds, plate, mean_blk_od)

#normalise od600 and set time to be in hours
df_all_analysis <- df_all_analysis %>%
  left_join(df_blanks) %>%
  mutate(norm_od = values - mean_blk_od) %>%
  filter(bacteria != "None") %>%
  mutate(Time_Hours = Time_seconds/(60*60)) %>%
  mutate(phage = factor(phage, levels = c("No Phage", "Tor", "PNM", "Cocktail"))) 

#calculate AUC for all control growth curves
control_auc_df <- df_all_analysis %>%
  filter(phage == "No Phage") %>%
  group_by(phage, bacteria, Well) %>%
  mutate(control_auc = DescTools::AUC(Time_Hours, norm_od, method = "trapezoid")) %>%
  group_by(bacteria) %>%
  reframe(mean_cntrl_auc = mean(control_auc)) %>%
  select(bacteria, mean_cntrl_auc)

#calculate virulence for each 
virulence_df <- df_all_analysis %>%
  group_by(phage, bacteria, Well) %>%
  mutate(auc = DescTools::AUC(Time_Hours, norm_od, method = "trapezoid")) %>%
  select(Well, phage, bacteria, auc) %>%
  unique() %>%
  left_join(control_auc_df) %>%
  mutate(vp = 1-(auc/mean_cntrl_auc)) %>%
  mutate(phage = factor(phage, levels = c("No Phage", "Tor", "PNM", "Cocktail")))

virulence_means <- virulence_df %>%
  group_by(phage, bacteria) %>%
  reframe(mean_vp = mean(vp), stdev = sd(vp))
  
virulence_means_plotting <- virulence_means %>%
  filter(phage != "No Phage") %>%
  mutate(does_infect = ifelse(mean_vp > 0.499, "Inhibition", "No Inhibition")) %>%
  mutate(does_infect = factor(does_infect, levels = c("Inhibition", "No Inhibition"))) %>%
  mutate(mean_vp = ifelse(mean_vp < 0, 0, mean_vp))

figure <- ggplot(virulence_means_plotting, aes(x = phage, y = bacteria, fill = mean_vp)) +
  geom_tile() +
  scale_fill_gradient2("Mean\nVirulence", high = "#a002a2", mid = "#f7d0f5", low = "white", midpoint = 0.6) +
  scale_y_discrete('Pseudomonas aeruginosa Strain') +
  scale_x_discrete("Phage") +
  labs(fill = "Inhibition") +
  theme_classic(base_size=28) +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_rect(colour = "black")) +
  theme(axis.text.x = element_text(angle = 90))

figure
ggsave("PDF-of-figures/figureS10.pdf", plot = figure, dpi = 300)
```

### Figure S11

```{r fig.asp=1, fig.width=12, warning=FALSE, message=FALSE, dpi=300, out.height='8.27in', out.width='8.27in'}
#| fig-cap: "For interactions between P.aeruginosa strain PAO1 and 10^7 PFU/mL of either phage Tor (n = 4), phage PNM (n = 4), or a 1:1 cocktail (n=2) of Tor and PNM where the mean virulence index is over 0.5, the mean virulence for each phage on each host is shown."
#| fig-dpi: 300

#read in the synogram data
row_content <- read.csv("rawdata/figureS11/row_data.csv")
col_content <- read.csv("rawdata/figureS11/column_data.csv")

#This function will reformat an individual .CSV into the correct format by modifying the columns, reformatting the time
#adding in the dictionary that tells us what is in each well, and removes NAs for dodgy values
reformat <- function(df_test) {
  df_test <- df_test %>% select(-Well)
  df_test[1,1] <- "Well"
  #names(df_test) <- lapply(df_test[1, ], as.character)
  df_test <- df_test[-1,] 
  df_test <- rename(df_test, "Time_Hours" = "X")
  
  df_test <- pivot_longer(df_test, cols = -Time_Hours, names_to = "Well", values_to = "values") %>%
    filter(Time_Hours != "Time")
  
  df_test <- df_test %>%
    mutate(
      hours = as.numeric(str_extract(Time_Hours, "\\d+(?=\\s*h)")),
      minutes = as.numeric(str_extract(Time_Hours, "\\d+(?=\\s*min)")),
      hours = replace_na(hours, 0),
      minutes = replace_na(minutes, 0),
      Time_seconds = (hours * 3600) + (minutes * 60)
    ) %>%
    select(-hours, -minutes)  # Remove temporary columns
  
  df_test <- df_test %>%
    mutate(
      Row = str_extract(Well, "^[A-Z]"),
      Column = as.numeric(str_extract(Well, "\\d+$"))
    )
}


#This function now will take each of the filenames in the folder, and for each it'll take in
#the file name, run the reformatting function above, add in a column with the plate name
#(taken from the file name) 

produce_df <- function(file_name){
  df <- read.csv(file_name, header = TRUE, row.names = NULL)
  df <- reformat(df)
  file_base_name <- sub("\\.\\w+$", "", basename(file_name))
  df$plate <- file_base_name
  return(df)
}

#get the list of all the csv files in the folder
csv_files <- list.files(
  path = "rawdata/figureS11/", 
  pattern = "\\.CSV$", 
  full.names = TRUE
)

#use purrr::map_df to apply the produce_df function to each CSV file and bind them together
df_all <- map(csv_files, produce_df)

#df in list form, this isnt ideal, need a concatenated one
df_list <- function(df_list){
  final_df <- Reduce(full_join, df_list)
  return(final_df)
}
df_all <- df_list(df_all)

df_all <- left_join(df_all, row_content) %>% left_join(col_content) %>% na.omit() %>%
  mutate(Row = factor(Row, levels = c("A","B","C","D","E","F","G"))) %>%
  mutate(Column = factor(Column, levels = seq(2,10,1)))

temporary_plot <- ggplot(df_all, aes(x = Time_seconds, y = values, group = Well)) +
  geom_line() +
  facet_grid(plate+Row~Column)

#data handling to make the rest of it make sense
df_all_analysis <- df_all %>%
  mutate(values = as.numeric(values)) %>%
  filter(Time_Hours != "")

#extract blanks
df_blanks <- df_all_analysis %>%
  filter(bacteria == "None") %>%
  filter(phage == "No Phage") %>%
  group_by(Time_seconds, plate) %>%
  mutate(mean_blk_od = mean(values)) %>%
  select(Time_seconds, plate, mean_blk_od)

#normalise od600 and set time to be in hours
df_all_analysis <- df_all_analysis %>%
  left_join(df_blanks) %>%
  mutate(norm_od = values - mean_blk_od) %>%
  filter(bacteria != "None") %>%
  mutate(Time_Hours = Time_seconds/(60*60)) %>%
  mutate(phage = factor(phage, levels = c("No Phage", "Tor", "PNM", "Cocktail"))) 


#calculate AUC for all control growth curves
control_auc_df <- df_all_analysis %>%
  filter(phage == "No Phage") %>%
  group_by(phage, bacteria, Well) %>%
  mutate(control_auc = DescTools::AUC(Time_Hours, norm_od, method = "trapezoid")) %>%
  group_by(bacteria) %>%
  reframe(mean_cntrl_auc = mean(control_auc)) %>%
  select(bacteria, mean_cntrl_auc)

#calculate virulence for each 
virulence_df <- df_all_analysis %>%
  group_by(phage, bacteria, Well) %>%
  mutate(auc = DescTools::AUC(Time_Hours, norm_od, method = "trapezoid")) %>%
  select(Well, phage, bacteria, auc) %>%
  unique() %>%
  left_join(control_auc_df) %>%
  mutate(vp = 1-(auc/mean_cntrl_auc)) %>%
  mutate(phage = factor(phage, levels = c("No Phage", "Tor", "PNM", "Cocktail")))

virulence_means <- virulence_df %>%
  group_by(phage, bacteria) %>%
  reframe(mean_vp = mean(vp), stdev = sd(vp))
  
virulence_means_plotting <- virulence_means %>%
  filter(phage != "No Phage") %>%
  mutate(does_infect = ifelse(mean_vp > 0.499, "Inhibition", "No Inhibition")) %>%
  mutate(does_infect = factor(does_infect, levels = c("Inhibition", "No Inhibition"))) %>%
  mutate(mean_vp = ifelse(mean_vp < 0, 0, mean_vp))

only_infect <- ggplot(filter(virulence_means_plotting, does_infect == "Inhibition"), aes(x = phage, y = mean_vp)) +
  geom_boxplot() +
  geom_point(position = position_dodge2(width = 0.4)) +
  scale_y_continuous("Mean Virulence Index", limits = c(0,1.5), breaks = seq(0,1.5,0.2)) +
  scale_x_discrete("Phage") +
  scale_colour_manual("Inhibtion\nover 48h", values = okabe) +
  theme_cowplot(28)

only_infect

ggsave("PDF-of-figures/figureS11.pdf", plot = only_infect, dpi = 300)
```

## Supplementary Text ST2

```{r}
#| label: session-info
#| echo: false

sessioninfo::package_info(pkgs = "loaded")
sessioninfo::session_info(to_file = "session.log")
```
